---
title: "R for Clinical Statistics: Wrangling II"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Build new columns with mutate, then summarise by group: the shape of most clinical reporting."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
```

## Welcome

Two new verbs this time, and they are the ones that turn a table of patients into the numbers you would put in a paper: `mutate`, which builds new columns, and `summarise`, which boils a column down to a single figure per group.

The same promises hold. You can't break anything, red text just means adjust something small, and **Hint** then **Solution** are always there.

## A quick warm-up

Start where M3 left off. **Run** this to remind yourself how `count()` works:

```{r recap, exercise=TRUE}
patients |>
  count(arm)
```

Two arms, split fairly evenly. Now let's make something new.

## Building columns with `mutate()`

`mutate()` adds a column worked out from the columns you already have. Every row stays; the table just gains a column. Run this, which flags everyone over 65:

```{r mutate-demo, exercise=TRUE}
patients |>
  mutate(over65 = age > 65)
```

Scroll to the right of the output and you'll find the new `over65` column, `TRUE` or `FALSE` for each patient.

Now your turn. A common clinical cut-off for obesity is a BMI of 30 and above. **Fill in the cut-off** so the new `bmi_cat` column reads `"Obese"` at or above it:

```{r mutate-bmicat, exercise=TRUE, exercise.blanks="___+"}
patients |>
  mutate(bmi_cat = if_else(bmi >= ____, "Obese", "Not obese"))
```

```{r mutate-bmicat-hint}
# A BMI of 30 or more counts as obese here.
# Type it as a plain number, with no quotes: 30
```

```{r mutate-bmicat-solution}
patients |>
  mutate(bmi_cat = if_else(bmi >= 30, "Obese", "Not obese"))
```

```{r mutate-bmicat-check}
grade_this({
  if (grepl('"30"', .user_code, fixed = TRUE) || grepl("'30'", .user_code, fixed = TRUE)) {
    fail("So close, and this one is sneaky: `\"30\"` in quotes is text, not a number. It happens to give the right answer for these BMI values, but text comparisons go character by character, so \"9\" counts as greater than \"30\". Drop the quotes and it is a real number.")
  }
  pass_if_equal(message = "Nicely done. Read `if_else` as a question with two answers: at or above 30, write Obese, otherwise Not obese.")
  fail("Not quite: the obesity cut-off is a BMI of 30, written as a plain number with no quotes.")
})
```

```{r mutate-bmicat-wrong, eval=FALSE, include=FALSE}
# expect: quotes
patients |>
  mutate(bmi_cat = if_else(bmi >= "30", "Obese", "Not obese"))
```

Keep `if_else(test, yes, no)` in mind: a question, then what to write when the answer is yes, then when it is no.

## Summarising by group

`summarise()` collapses a whole column into one number. On its own it gives a single figure for the entire trial. Run this for the mean systolic blood pressure:

```{r summarise-demo, exercise=TRUE}
patients |>
  summarise(mean_sbp = mean(sbp))
```

One row, one number. Useful, though you usually want it *per arm*, to compare the treatment groups. That is what `group_by()` adds.

**Fill in the column to group by** so you get a separate mean for each treatment arm:

```{r groupby-arm, exercise=TRUE, exercise.blanks="___+"}
patients |>
  group_by(____) |>
  summarise(mean_sbp = mean(sbp))
```

```{r groupby-arm-hint}
# You want one mean per treatment arm.
# The column is called: arm
```

```{r groupby-arm-solution}
patients |>
  group_by(arm) |>
  summarise(mean_sbp = mean(sbp))
```

```{r groupby-arm-check}
grade_this({
  pass_if_equal(message = "There it is: a mean blood pressure for each arm, side by side. Swap `arm` for `smoker` and you would compare smokers instead.")
  fail("Almost: group by `arm` so summarise gives one row per treatment group.")
})
```

The summary doesn't have to be a mean. **Fill in the function** to average the SBP in each arm (the average is the `mean`):

```{r summarise-fn, exercise=TRUE, exercise.blanks="___+"}
patients |>
  group_by(arm) |>
  summarise(mean_sbp = ____(sbp))
```

```{r summarise-fn-hint}
# The function for an average is called: mean
```

```{r summarise-fn-solution}
patients |>
  group_by(arm) |>
  summarise(mean_sbp = mean(sbp))
```

```{r summarise-fn-check}
grade_this({
  pass_if_equal(message = "Exactly. `mean()` is one of many: `median(sbp)`, `min(sbp)` and `max(sbp)` work the same way.")
  fail("Not quite: the function for an average is `mean`.")
})
```

## Practice

Your turn, no blanks. Write each one out. The **Hint** and **Solution** are a click away.

**1.** Add a logical column called `high_bp` that is `TRUE` when `sbp` is above 140.

```{r pr-mutate, exercise=TRUE}
# Your code here

```

```{r pr-mutate-hint}
# mutate() adds a column. A comparison gives TRUE/FALSE:
# patients |> mutate(high_bp = sbp > 140)
```

```{r pr-mutate-solution}
patients |>
  mutate(high_bp = sbp > 140)
```

```{r pr-mutate-check}
grade_this({
  pass_if_equal(message = "Done. `high_bp` is TRUE wherever systolic pressure clears 140, one answer per patient.")
  fail("Not quite. Add a column with `mutate(high_bp = sbp > 140)`.")
})
```

**2.** For each `smoker` group, find the mean BMI. Call the summary column `mean_bmi`.

```{r pr-groupsum, exercise=TRUE}
# Your code here

```

```{r pr-groupsum-hint}
# Group first, then summarise:
# patients |> group_by(smoker) |> summarise(mean_bmi = mean(bmi))
```

```{r pr-groupsum-solution}
patients |>
  group_by(smoker) |>
  summarise(mean_bmi = mean(bmi))
```

```{r pr-groupsum-check}
grade_this({
  pass_if_equal(message = "One mean BMI per smoking group, side by side. Group first, then summarise.")
  fail("Close. Group by `smoker`, then `summarise(mean_bmi = mean(bmi))`. Keep the name `mean_bmi` so it matches.")
})
```

**3.** Among patients on the `Active` arm only, find the mean `sbp` for each `smoker` group. Call it `mean_sbp`.

```{r pr-combo, exercise=TRUE}
# Your code here

```

```{r pr-combo-hint}
# Three steps: filter the arm, group by smoker, then summarise.
# patients |> filter(arm == "Active") |> group_by(smoker) |> summarise(mean_sbp = mean(sbp))
```

```{r pr-combo-solution}
patients |>
  filter(arm == "Active") |>
  group_by(smoker) |>
  summarise(mean_sbp = mean(sbp))
```

```{r pr-combo-check}
grade_this({
  pass_if_equal(message = "That brings three verbs together: filter the arm, group by smoker, summarise the mean. Most analyses look like this.")
  fail("Almost. Filter to the Active arm, then group by `smoker`, then `summarise(mean_sbp = mean(sbp))`.")
})
```

## Recap

Two verbs that do most of the work of reporting:

- **`mutate()`** builds a new column from existing ones
- **`group_by()`** with **`summarise()`** reduces a column to one number per group

Put these together with M3's `filter`, `count`, `arrange` and `select`, and you have the core of day-to-day analysis. Most of the "Table 1" you see in a clinical paper is a handful of lines like these.
