---
title: "R for Clinical Statistics: Reproducible Reports"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "The capstone: rebuild the whole trial analysis as a Quarto report that renders to a document, with every number computed from the data."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(survival)
library(finalfit)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")
data("outcomes", package = "r4clinstats")

bp_change <- labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value, names_prefix = "v") |>
  left_join(select(patients, id, arm), by = "id") |>
  mutate(
    change = v2 - v0,
    arm = relevel(factor(arm), ref = "Placebo")
  ) |>
  select(id, arm, baseline = v0, week24 = v2, change)

bp_mean <- labs |>
  filter(measure == "sbp") |>
  left_join(select(patients, id, arm), by = "id") |>
  group_by(arm, weeks) |>
  summarise(sbp = mean(value), .groups = "drop")

trial <- outcomes |>
  left_join(select(patients, id, age, sex, arm), by = "id") |>
  mutate(
    arm = relevel(factor(arm), ref = "Placebo"),
    response = factor(response)
  )

km <- survfit(Surv(time, status) ~ arm, data = trial)
```

## Welcome to the last lesson

In the very first lesson, I made a claim: *the script is your method section*. A written analysis that anyone can read, re-run and check. Today you cash that claim in.

This lesson ends with something no other lesson has produced: a real document on your own computer. You will rebuild this course's entire trial analysis — the Table 1, the blood-pressure model, the odds ratio, the survival curve — inside one **Quarto** file, press one button, and hold the report it builds.

Two things are different about this final lesson. First, there are **no blanks anywhere**: every exercise asks you to write the whole thing, because after thirteen lessons you can. Second, the lesson does not end in this tutorial. It ends in RStudio, with a rendered report. The promises still hold the whole way: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away.

## What a Quarto document is

A Quarto file (`.qmd`) is plain text with three kinds of content. At the top, a **YAML header** describing the document:

````
---
title: "Systolic blood pressure trial: analysis report"
author: "Your name here"
date: today
format: html
---
````

Then ordinary **prose**, written in the same lightweight formatting you have seen everywhere (`## Heading`, `**bold**`). And between the paragraphs, **code chunks**: the same R you have written all course, fenced off so Quarto knows to run it.

````
```{r}
#| label: table-one
patients |>
  summary_factorlist("arm", c("age", "sex", "smoker", "bmi"))
```
````

When you press **Render**, Quarto runs every chunk from top to bottom in a fresh session and weaves the output — tables, figures, numbers — into a finished document: HTML, Word, or PDF. The document *is* the analysis. There is no copy-paste step where a stale number can sneak in, because there is no copying.

One habit before anything else: **render early, render often.** A document that rendered five minutes ago and breaks now leaves only one suspect.

## Numbers that write themselves

The quiet superpower is **inline code**: a fragment of R inside a sentence. In the prose of a `.qmd` you can write:

```
The trial randomised `r nrow(patients)` patients.
```

and the rendered document says "The trial randomised 100 patients." If the data change, the sentence updates itself on the next render. A results paragraph written this way can never disagree with its own tables.

**Run this** to see the kind of expression that lives inside those backticks:

```{r inline-demo, exercise=TRUE}
paste("The trial randomised", nrow(patients), "patients.")
```

Small expressions like `nrow(patients)` or `round(estimate, 1)` are all inline code ever needs to be. Anything bigger belongs in a chunk.

## The chunks of your report

Time to build. The six exercises below are the six analysis chunks of your report, and together they are the whole course in miniature. Everything you need is loaded: `patients`, `bp_change`, `bp_mean`, `trial`, and a fitted Kaplan-Meier estimate `km`, all built exactly as in Lessons 11 to 13 (with `Placebo` as the reference level).

### The baseline table

Every report opens by describing who was in it. Build the Table 1: `age`, `sex`, `smoker` and `bmi`, compared across `arm`, from `patients`.

```{r rep-table, exercise=TRUE}
# Your code here

```

```{r rep-table-hint}
# Lesson 9's one-liner:
# patients |> summary_factorlist("arm", c("age", "sex", "smoker", "bmi"))
```

```{r rep-table-solution}
patients |>
  summary_factorlist("arm", c("age", "sex", "smoker", "bmi"))
```

```{r rep-table-check}
grade_this({
  pass_if_equal(message = "Chunk one of your report, done. In the report itself you will pipe this into `kable()`, which turns a raw data frame into a clean document table.")
  if (is.data.frame(.result) && "label" %in% names(.result)) {
    vars <- setdiff(unique(.result$label), "")
    if (!"Active" %in% names(.result)) {
      fail("The columns should be the two treatment arms, so the dependent is `\"arm\"`.")
    }
    if (!setequal(vars, c("age", "sex", "smoker", "bmi"))) {
      fail("Check the variable list: this table describes `age`, `sex`, `smoker` and `bmi`.")
    }
  }
  fail("Not quite: this is Lesson 9's `summary_factorlist(\"arm\", c(\"age\", \"sex\", \"smoker\", \"bmi\"))`, on `patients`.")
})
```

```{r rep-table-wrong, eval=FALSE, include=FALSE}
# expect: variable list
patients |>
  summary_factorlist("arm", c("age", "sex"))
```

### The trend figure

The figure that shows the whole trial at a glance: from `bp_mean`, the mean blood pressure over `weeks`, one coloured line per `arm`, with a readable y-axis label from `labs()`. Polish it however you like; it is going in a report.

```{r rep-trend, exercise=TRUE}
# Your code here

```

```{r rep-trend-hint}
# Lesson 8's polished trend:
# ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
#   geom_line(linewidth = 1) +
#   labs(y = "Mean systolic BP (mmHg)") +
#   theme_minimal()
```

```{r rep-trend-solution}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
  geom_line(linewidth = 1) +
  labs(y = "Mean systolic BP (mmHg)") +
  theme_minimal()
```

```{r rep-trend-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start with `ggplot(bp_mean, aes(...))`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomLine" %in% geoms)) {
    fail("Draw the trend with `geom_line()`.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if (!("arm" %in% cols)) {
    fail("Map `arm` to `colour` so each arm gets its own line.")
  }
  ylab <- .result$labels$y
  if (is.null(ylab) || ylab %in% c("sbp", "value")) {
    fail("Give the y-axis a readable label with `labs(y = \"...\")` — a figure in a report never shows a raw column name.")
  }
  pass("A figure fit for a paper. When this chunk runs in your report, the figure lands on the page right where the chunk sits.")
})
```

### The primary analysis

The trial's headline model, from Lesson 12: `change` modelled on `baseline` and `arm` in `bp_change`, tidied with a confidence interval.

```{r rep-model, exercise=TRUE}
# Your code here

```

```{r rep-model-hint}
# lm(change ~ baseline + arm, data = bp_change) |>
#   tidy(conf.int = TRUE)
```

```{r rep-model-solution}
lm(change ~ baseline + arm, data = bp_change) |>
  tidy(conf.int = TRUE)
```

```{r rep-model-check}
grade_this({
  pass_if_equal(message = "The baseline-adjusted treatment effect: 5.0 mmHg further fall, p = 0.002. In the report, the sentence underneath will quote these numbers with inline code, so it can never drift from this table.")
  if (inherits(.result, "lm")) {
    fail("The model is fitted, but that's the raw model object. Pipe it into `tidy(conf.int = TRUE)`.")
  }
  if (is.data.frame(.result) && "term" %in% names(.result)) {
    if (!"baseline" %in% .result$term) {
      fail("This one is the baseline-adjusted model, so the formula is `change ~ baseline + arm`.")
    }
    if (!"conf.low" %in% names(.result)) {
      fail("The estimates are there but the confidence interval isn't. Add `conf.int = TRUE` inside `tidy()`.")
    }
  }
  fail("Not quite: `lm(change ~ baseline + arm, data = bp_change)`, then `tidy(conf.int = TRUE)`.")
})
```

```{r rep-model-wrong, eval=FALSE, include=FALSE}
# expect: confidence interval
lm(change ~ baseline + arm, data = bp_change) |>
  tidy()
```

### The odds ratio

The response analysis: logistic regression of `response` on `arm` in `trial`, tidied into an odds ratio with its confidence interval.

```{r rep-or, exercise=TRUE}
# Your code here

```

```{r rep-or-hint}
# glm(response ~ arm, data = trial, family = binomial) |>
#   tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r rep-or-solution}
glm(response ~ arm, data = trial, family = binomial) |>
  tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r rep-or-check}
grade_this({
  pass_if_equal(message = "OR 2.66, CI 1.17 to 6.30. Chunk four of six.")
  if (inherits(.result, "glm")) {
    fail("The model is fitted, but that's the raw model object. Pipe it into `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
  }
  if (is.data.frame(.result) && "term" %in% names(.result)) {
    if (!"conf.low" %in% names(.result)) {
      fail("The odds ratio is there but the confidence interval isn't. Add `conf.int = TRUE` inside `tidy()`.")
    }
    if (length(.result$estimate) == length(.solution$estimate) &&
        isTRUE(all.equal(.result$estimate, log(.solution$estimate)))) {
      fail("The model is right, but those are log-odds rather than odds ratios. Add `exponentiate = TRUE` inside `tidy()`.")
    }
  }
  fail("Not quite: `glm(response ~ arm, data = trial, family = binomial)`, then `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
})
```

```{r rep-or-wrong, eval=FALSE, include=FALSE}
# expect: log-odds
glm(response ~ arm, data = trial, family = binomial) |>
  tidy(conf.int = TRUE)
```

### The survival curve

The Kaplan-Meier staircases from Lesson 13: `km` is already fitted; tidy it and draw `estimate` against `time`, one coloured step per `strata`.

```{r rep-km, exercise=TRUE}
# Your code here

```

```{r rep-km-hint}
# tidy(km) |>
#   ggplot(aes(x = time, y = estimate, colour = strata)) +
#   geom_step()
```

```{r rep-km-solution}
tidy(km) |>
  ggplot(aes(x = time, y = estimate, colour = strata)) +
  geom_step()
```

```{r rep-km-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start from `tidy(km)`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomStep" %in% geoms)) {
    fail("Survival curves are staircases: `geom_step()`.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if (!("strata" %in% cols)) {
    fail("Map `strata` to `colour` so each arm gets its own staircase.")
  }
  pass("The figure every trial report saves for last: two staircases parting over the year. Add `labs()` in the report if you want it fully dressed.")
})
```

### The hazard ratio

And the number that goes with it: the Cox model of the treatment effect in `trial`, tidied into a hazard ratio with its confidence interval.

```{r rep-cox, exercise=TRUE}
# Your code here

```

```{r rep-cox-hint}
# coxph(Surv(time, status) ~ arm, data = trial) |>
#   tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r rep-cox-solution}
coxph(Surv(time, status) ~ arm, data = trial) |>
  tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r rep-cox-check}
grade_this({
  pass_if_equal(message = "HR 0.47, CI 0.25 to 0.88. That is all six chunks: the whole analysis, rebuilt from memory. Now you get to put them somewhere permanent.")
  if (inherits(.result, "coxph")) {
    fail("The model is fitted, but that's the raw model object. Pipe it into `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
  }
  if (is.data.frame(.result) && "term" %in% names(.result)) {
    if (!"conf.low" %in% names(.result)) {
      fail("The hazard ratio is there but the confidence interval isn't. Add `conf.int = TRUE` inside `tidy()`.")
    }
    if (length(.result$estimate) == length(.solution$estimate) &&
        isTRUE(all.equal(.result$estimate, log(.solution$estimate)))) {
      fail("The model is right, but those are log hazard ratios. Add `exponentiate = TRUE` inside `tidy()`.")
    }
  }
  fail("Not quite: `coxph(Surv(time, status) ~ arm, data = trial)`, then `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
})
```

```{r rep-cox-wrong, eval=FALSE, include=FALSE}
# expect: raw model object
coxph(Surv(time, status) ~ arm, data = trial)
```

## Now build the real thing

Here is where you leave the sandbox. Over in **RStudio's console** (not in this tutorial), run:

```
r4clinstats::start_report()
```

It puts a file called `trial-report.qmd` in your working directory and opens it. Look around: the YAML header, the prose, the setup chunk that builds `bp_change` and `trial` exactly as this lesson did, and six TODO chunks with the same names as the six exercises you just passed.

Then work like this:

1. **Press Render first**, before touching anything. The document builds as it stands: your first reproducible report, thirty seconds in. (RStudio ships with Quarto built in — if you can see the Render button, you have everything.)
2. **Fill one TODO chunk**, using your code from the matching exercise above, and render again. The section appears. One difference from the sandbox: pipe each table into `kable()`, which typesets a data frame properly for a document.
3. **Repeat** until all six are in. Then write the conclusion in your own words, quoting numbers with inline code the way the Patients paragraph already does.

Two finishing touches worth trying. Put your actual name in the YAML `author` field and render: your name, on your analysis. And change `format: html` to `format: docx`, render once more, and Quarto hands you a Word document — the same analysis, in the format half of clinical research still runs on.

If you get stuck, `start_report(completed = TRUE, path = "trial-report-completed.qmd")` gives you the finished report to compare against, and every exercise above has your answers.

## Taking this to your own data

Your next report will not use course data, so here is the shape of a real project, in five lines:

1. In RStudio: **File ▸ New Project ▸ New Directory**. A project keeps one analysis in one folder.
2. Make a `data/` folder inside it and put your raw file there. **Never edit the raw file again** (Lesson 6's rule).
3. Start a `.qmd`, and in its setup chunk read the data with `read_csv(here("data", "yourfile.csv"))`.
4. Clean, describe, analyse — chunk by chunk, rendering as you go. Every tool you need is in the thirteen lessons behind you.
5. When a collaborator asks "what exactly did you do?", send the `.qmd`. That is the whole answer.

And when you outgrow the course, four books are worth your time. They were introduced back in Lesson 0, and they are the natural next step now:

- **[HealthyR: R for Health Data Science](https://argoshare.is.ed.ac.uk/healthyr_book/)** — the closest companion to this course, and home ground for `finalfit`.
- **[R for Data Science](https://r4ds.hadley.nz/)** — the tidyverse, thoroughly.
- **[The Epidemiologist R Handbook](https://epirhandbook.com/)** — cleaning and analysing real health data in the wild.
- **[R Graphics Cookbook](https://r-graphics.org/)** — the book to open when you want a particular plot.

## The end

A recap of the lesson, and of everything:

- a **Quarto document** holds prose, code and results in one plain-text file; **Render** rebuilds the whole thing from the data
- **inline code** writes the numbers into your sentences, so the text can never disagree with the tables
- your report's six chunks were Lessons 9 to 13, written from memory, with no blanks

Fourteen lessons ago you were promised you could not break anything, and you never did. You went from one line of R to a randomised trial analysed end to end — designed tidy, wrangled, cleaned, described, pictured, tested, modelled, survived, and now written up in a document that rebuilds itself on demand.

That last part was always the point. The script is your method section. Go write one about something real.
