---
title: "R for Clinical Statistics: Survival Analysis"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Time-to-event data done properly: censoring, Kaplan-Meier curves, the log-rank test, and hazard ratios from a Cox model."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)
library(broom)
library(survival)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("outcomes", package = "r4clinstats")

# Outcomes joined to patient details; Placebo as the reference level so
# hazard ratios read as the effect of the active drug.
trial <- outcomes |>
  left_join(select(patients, id, age, sex, arm), by = "id") |>
  mutate(arm = relevel(factor(arm), ref = "Placebo"))
```

## Welcome

The most clinical question of all is *how long*: how long until relapse, readmission, recovery, death. Time-to-event data has a twist that breaks every tool you have met so far, and this lesson gives you the tools built for it: the **Kaplan-Meier curve**, the **log-rank test**, and the **Cox model** with its hazard ratios.

The promises hold: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away.

## The twist: censoring

Look at the trial data again. **Run this:**

```{r meet-trial, exercise=TRUE}
glimpse(trial)
```

Each patient has a `time` in days and a `status`: 1 if the event happened then, 0 if the year ended first. That 0 is called **censoring**, and 58 patients have it. A censored patient is not missing data; they carry real information — they lasted *at least* that long — and simply averaging the times would throw it away (the describing-data lesson hinted at this). Survival methods exist to use both numbers honestly.

R packages both into one object with `Surv()`. **Run this** to see the first ten patients:

```{r surv-demo, exercise=TRUE}
head(Surv(trial$time, trial$status), 10)
```

Plain numbers are events; numbers wearing a `+` are censored, read as "365-plus: still event-free when we stopped watching."

## The Kaplan-Meier curve

`survfit()` estimates the survival curve: the proportion still event-free at each point in time. The formula reads like every other model in this course. **Fill in the grouping column** to estimate one curve per treatment arm:

```{r km, exercise=TRUE, exercise.blanks="___+"}
km <- survfit(Surv(time, status) ~ ____, data = trial)
km
```

```{r km-hint}
# One curve per treatment group, so the right-hand side is:
# arm
```

```{r km-solution}
km <- survfit(Surv(time, status) ~ arm, data = trial)
km
```

```{r km-check}
grade_this({
  if (!inherits(.result, "survfit")) {
    fail("That should build a survival estimate with `survfit(...)`. Keep the structure and fill in the grouping column.")
  }
  if (identical(sort(names(.result$strata)), c("arm=Active", "arm=Placebo"))) {
    pass("Read the table: 27 events among 51 placebo patients with a median of 341 days, but only 15 events among 49 on the drug, with median `NA`. That NA is good news, and the next paragraph decodes it.")
  }
  fail("Almost: the formula's right-hand side is `arm`, one curve per treatment group.")
})
```

```{r km-wrong, eval=FALSE, include=FALSE}
# expect: arm
km <- survfit(Surv(time, status) ~ sex, data = trial)
km
```

The **median survival** is the day the curve crosses 50%. The active arm's median is `NA` because its curve never fell that far: more than half the treated patients were still event-free at one year. Papers write this as *"median not reached"*, and it is one of the most encouraging sentences a trial can print.

## Drawing the curves

A Kaplan-Meier curve is a staircase: flat while nothing happens, stepping down at each event. `tidy()` turns the fit into a plottable table, and the staircase needs the right geom. **Fill it in:**

```{r km-plot, exercise=TRUE, exercise.blanks="___+"}
tidy(km) |>
  ggplot(aes(x = time, y = estimate, colour = strata)) +
  ____()
```

```{r km-plot-setup}
km <- survfit(Surv(time, status) ~ arm, data = trial)
```

```{r km-plot-hint}
# Survival curves step down at each event; the geom that draws steps is:
# geom_step
```

```{r km-plot-solution}
tidy(km) |>
  ggplot(aes(x = time, y = estimate, colour = strata)) +
  geom_step()
```

```{r km-plot-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomStep" %in% geoms) {
    pass("The classic trial figure: two staircases separating over the year, with the active arm staying higher. Every oncology paper you have ever read opens with this picture.")
  }
  fail("Almost: a survival curve is a staircase, drawn with `geom_step()`.")
})
```

## Testing the gap: the log-rank test

The curves separate, and the **log-rank test** asks whether that separation outruns chance. It compares the events each group actually had with the number it would expect if the curves were truly the same. **Run it:**

```{r logrank, exercise=TRUE}
survdiff(Surv(time, status) ~ arm, data = trial)
```

The placebo arm had 27 events where 19 were expected; the active arm 15 where 23 were expected. p = 0.02: the separation is real.

## The hazard ratio: a Cox model

The log-rank test says *whether* the curves differ; the **Cox model** says *by how much*, and lets you adjust for other variables, exactly as `lm()` and `glm()` did. Its summary is the **hazard ratio** (HR): the event rate on the drug relative to placebo, where below 1 means protection. **Run the model**, then tidy it the way you tidied the odds ratios — **fill in the value:**

```{r cox, exercise=TRUE, exercise.blanks="___+"}
cox <- coxph(Surv(time, status) ~ arm, data = trial)
tidy(cox, exponentiate = ____, conf.int = TRUE)
```

```{r cox-hint}
# Same move as the odds ratios: exponentiate the coefficients.
# The value is: TRUE
```

```{r cox-solution}
cox <- coxph(Surv(time, status) ~ arm, data = trial)
tidy(cox, exponentiate = TRUE, conf.int = TRUE)
```

```{r cox-check}
grade_this({
  pass_if_equal(message = "HR 0.47, CI 0.25 to 0.88, p = 0.018: the drug roughly halves the event rate, and the interval sits clear of 1. That one row is the trial's survival story.")
  fail("Almost: `exponentiate = TRUE` turns the coefficients into hazard ratios.")
})
```

## Practice

Three to finish, no blanks. Write each one yourself; **Hint** and **Solution** are there if you want them.

**1.** Estimate Kaplan-Meier curves split by `sex` instead of arm.

```{r pr-km, exercise=TRUE}
# Your code here

```

```{r pr-km-hint}
# survfit(Surv(time, status) ~ sex, data = trial)
```

```{r pr-km-solution}
survfit(Surv(time, status) ~ sex, data = trial)
```

```{r pr-km-check}
grade_this({
  if (!inherits(.result, "survfit")) {
    fail("Build the estimate with `survfit(Surv(time, status) ~ ..., data = trial)`.")
  }
  if (identical(sort(names(.result$strata)), c("sex=Female", "sex=Male"))) {
    pass("Neither sex reaches its median and the event counts sit close to expectation: no sign survival differs by sex. A null result, correctly obtained.")
  }
  fail("Close. The right-hand side of the formula should be `sex`.")
})
```

**2.** Draw the treatment-arm staircases from scratch: tidy `km` (it is already built for you), then plot `estimate` against `time`, one coloured step line per `strata`.

```{r pr-plot, exercise=TRUE}
# Your code here

```

```{r pr-plot-setup}
km <- survfit(Surv(time, status) ~ arm, data = trial)
```

```{r pr-plot-hint}
# tidy(km) |>
#   ggplot(aes(x = time, y = estimate, colour = strata)) +
#   geom_step()
```

```{r pr-plot-solution}
tidy(km) |>
  ggplot(aes(x = time, y = estimate, colour = strata)) +
  geom_step()
```

```{r pr-plot-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start from `tidy(km)`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomStep" %in% geoms)) {
    fail("The staircase geom is `geom_step()`.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if (!("strata" %in% cols)) {
    fail("Map `strata` to `colour` so each arm gets its own staircase.")
  }
  pass("A publication figure from first principles: tidy the fit, map the aesthetics, step the line.")
})
```

**3.** Fit a Cox model of the treatment effect adjusted for `age`, and tidy it into hazard ratios with confidence intervals.

```{r pr-cox, exercise=TRUE}
# Your code here

```

```{r pr-cox-hint}
# coxph(Surv(time, status) ~ arm + age, data = trial) |>
#   tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r pr-cox-solution}
coxph(Surv(time, status) ~ arm + age, data = trial) |>
  tidy(exponentiate = TRUE, conf.int = TRUE)
```

```{r pr-cox-check}
grade_this({
  pass_if_equal(message = "Age-adjusted HR 0.48, CI 0.26 to 0.91: the effect barely moves, so age was not doing the work. Adjustment in a Cox model is the same `+ age` it was everywhere else.")
  if (inherits(.result, "coxph")) {
    fail("The model is fitted, but that's the raw model object. Pipe it into `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
  }
  if (is.data.frame(.result) && "term" %in% names(.result)) {
    if (!"age" %in% .result$term) {
      fail("Age isn't in the model yet. Adjusting for it means `Surv(time, status) ~ arm + age`.")
    }
    if (!"conf.low" %in% names(.result)) {
      fail("The hazard ratios are there but the confidence interval isn't. Add `conf.int = TRUE` inside `tidy()`.")
    }
    if (length(.result$estimate) == length(.solution$estimate) &&
        isTRUE(all.equal(.result$estimate, log(.solution$estimate)))) {
      fail("The model is right, but those are log hazard ratios. Add `exponentiate = TRUE` inside `tidy()`.")
    }
  }
  fail("Not quite: `coxph(Surv(time, status) ~ arm + age, data = trial)`, then `tidy(exponentiate = TRUE, conf.int = TRUE)`.")
})
```

```{r pr-cox-wrong, eval=FALSE, include=FALSE}
# expect: raw model object
coxph(Surv(time, status) ~ arm + age, data = trial)
```

## Recap

Time-to-event analysis in four moves:

- **`Surv(time, status)`** packages follow-up and censoring into one outcome
- **`survfit()`** estimates the Kaplan-Meier curve; *median not reached* means the curve never fell below half
- **`survdiff()`** runs the log-rank test for whether curves differ
- **`coxph()`** puts a size on it: the **hazard ratio**, tidied with `exponentiate = TRUE`

That completes the analysis toolkit. You can now take a trial from a raw spreadsheet to a survival curve and a hazard ratio: design the data, wrangle it, describe it, picture it, and test what it shows — every step written down, repeatable, and yours.

One lesson remains, and it is the payoff for all of this: turning the whole analysis into a document you can hand to a colleague.
