---
title: "R for Clinical Statistics: Comparing Groups"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Your first statistical tests: the t-test for numbers, the chi-squared test for categories, and tidy results with broom."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")
data("outcomes", package = "r4clinstats")

# One row per patient: baseline SBP, week-24 SBP, and the change between them.
bp_change <- labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value, names_prefix = "v") |>
  left_join(select(patients, id, arm), by = "id") |>
  mutate(change = v2 - v0) |>
  select(id, arm, baseline = v0, week24 = v2, change)

# Outcomes with each patient's details attached.
trial <- outcomes |>
  left_join(select(patients, id, age, sex, arm), by = "id")
```

## Welcome

You can describe a difference; now you test one. Two groups differ a little in any sample, by chance alone. A statistical test asks: **is this gap bigger than chance could plausibly produce?** This lesson covers the two tests that answer most clinical questions: the **t-test** for a number, and the **chi-squared test** for categories.

The promises hold: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away.

## Meet the change scores

The setup for this lesson quietly did some familiar work: it took the `labs` blood pressures, pivoted them wide (one row per patient, a column per visit), and computed each patient's **change** from baseline to week 24. The result is `bp_change`. **Run this:**

```{r meet-change, exercise=TRUE}
glimpse(bp_change)
```

One row per patient: their arm, their `baseline` blood pressure, their `week24` reading, and the `change` between the two. A negative change means blood pressure fell. This table is where the trial's real question lives.

## The question, drawn first

Before any test, look. **Run this** box plot of the changes:

```{r box-change, exercise=TRUE}
ggplot(bp_change, aes(x = arm, y = change)) +
  geom_boxplot()
```

The active arm's box sits lower: blood pressure fell more on the drug. The boxes overlap plenty, though. Chance, or effect? That is precisely a t-test's job.

## The t-test

`t.test()` compares the mean of a number between two groups, and it wants the comparison written as a **formula**: `change ~ arm`, which reads *"change, split by arm."*

That `~` is a **tilde**, and it is worth a moment because it runs through the rest of the course. It is not a typo and not a minus sign. R uses it to mean "by", or "explained by": what you are measuring goes on the left, and what splits or explains it goes on the right. You have already met it in Lesson 8 as `facet_wrap(~ arm)`, meaning one panel *by* arm. Every test and model from here on is written the same way.

First, though, a deliberately boring example: comparing the arms' **baseline** blood pressure. **Run it:**

```{r ttest-baseline, exercise=TRUE}
t.test(sbp ~ arm, data = patients)
```

The two means sit within half a point of each other and p = 0.91: no evidence of any difference, exactly as it should be, because randomisation built the arms to start equal. (This is why a trial's baseline table carries no p-values, the point from the tables lesson.)

The real question is whether the arms *changed* differently. **Fill in the grouping column:**

```{r ttest-change, exercise=TRUE, exercise.blanks="___+"}
t.test(change ~ ____, data = bp_change)
```

```{r ttest-change-hint}
# Split the changes by treatment group. The column in bp_change is:
# arm
```

```{r ttest-change-solution}
t.test(change ~ arm, data = bp_change)
```

```{r ttest-change-check}
grade_this({
  if (!inherits(.result, "htest")) {
    fail("That should run a test. Keep the `t.test(...)` structure and fill in the grouping column.")
  }
  if (isTRUE(all.equal(.result$p.value, .solution$p.value))) {
    pass("There is the answer: a mean fall of 5.0 mmHg on the active arm against 0.1 on placebo, p = 0.039. A gap that size would arise by chance less than one time in twenty five.")
  }
  fail("Almost: the grouping column is `arm`. Read the formula as: change, split by arm.")
})
```

Three lines of that output earn a place in a paper: the two group means, the confidence interval for the difference (here, the active arm fell 0.2 to 9.4 mmHg further), and the p-value.

## The chi-squared test

When both variables are categories, the tool is the **chi-squared test**. The trial's other outcome is `response`, a plain Y or N, and it lives in the `trial` table. Look at the raw counts first. **Run this:**

```{r count-response, exercise=TRUE}
trial |>
  count(arm, response)
```

36 of 49 responded on the active arm (73%) against 26 of 51 on placebo (51%). The test takes those counts as a table. **Fill in the second column:**

```{r chisq, exercise=TRUE, exercise.blanks="___+"}
chisq.test(table(trial$arm, trial$____))
```

```{r chisq-hint}
# You are testing whether response depends on arm.
# The second column is: response
```

```{r chisq-solution}
chisq.test(table(trial$arm, trial$response))
```

```{r chisq-check}
grade_this({
  if (!inherits(.result, "htest")) {
    fail("That should run a test. Keep the `chisq.test(table(...))` structure.")
  }
  if (isTRUE(all.equal(.result$p.value, .solution$p.value))) {
    pass("p = 0.035: response and arm are associated beyond what chance comfortably explains. 73% against 51% is a difference worth having.")
  }
  fail("Almost: the second column in the table should be `response`.")
})
```

```{r chisq-wrong, eval=FALSE, include=FALSE}
# expect: response
chisq.test(table(trial$arm, trial$adverse_event))
```

## Tidy results with broom

Test output is written for reading, not for reuse. `t.test()` prints a tidy little paragraph, and there is no easy way to pull the p-value out of a paragraph and into a table.

**broom** is a small package that exists to solve exactly that. It takes the output of a test or a model and hands it back as a data frame, with plain column names like `estimate`, `conf.low` and `p.value`. Its main function is `tidy()`, and the same call works on a t-test, a chi-squared, a linear model or a survival model, which is why it appears in every lesson from here to the end. Once a result is a data frame, every verb from Lessons 3 and 4 works on it.

**Fill in the function:**

```{r tidy-test, exercise=TRUE, exercise.blanks="___+"}
t.test(change ~ arm, data = bp_change) |>
  ____()
```

```{r tidy-test-hint}
# The broom function that turns a test into a tidy data frame is:
# tidy
```

```{r tidy-test-solution}
t.test(change ~ arm, data = bp_change) |>
  tidy()
```

```{r tidy-test-check}
grade_this({
  pass_if_equal(message = "The whole test on one tidy row: estimates, confidence interval, p-value. Every test and model in the rest of the course can be tidied the same way.")
  fail("Almost: pipe the test into `tidy()`, from the broom package.")
})
```

### The columns you cannot see

Look at the top right of that result. It says something like **1-4 of 10 columns**, with a small arrow beside it.

`tidy()` hands back *everything* the test produced, and for a t-test that is ten columns. The output panel is only so wide, so it shows the first few and hides the rest behind that arrow. The confidence interval and the p-value are among the hidden ones, which is worth knowing before you conclude they are missing. Click the arrow to page across.

Hunting through ten columns gets old. Since the result is now an ordinary data frame, `select()` from Lesson 3 works on it, and you can keep just the columns you would actually report. **Run this:**

```{r tidy-select, exercise=TRUE}
t.test(change ~ arm, data = bp_change) |>
  tidy() |>
  select(estimate, conf.low, conf.high, p.value)
```

Four columns, all visible at once: the difference between the arms, the interval around it, and the p-value. That is the whole result of the trial, and it is the habit to carry into the modelling lessons, where the tables get wider still.

## Practice

Three to finish, no blanks. Write each one yourself; **Hint** and **Solution** are there if you want them.

**1.** Test whether `bmi` differs between the treatment arms in `patients`.

```{r pr-ttest, exercise=TRUE}
# Your code here

```

```{r pr-ttest-hint-1}
# Formula syntax: the number you are measuring, a tilde, then the column
# that splits it into two groups. The table goes in the data argument.
```

```{r pr-ttest-hint-2}
# t.test(bmi ~ arm, data = patients)
```

```{r pr-ttest-solution}
t.test(bmi ~ arm, data = patients)
```

```{r pr-ttest-check}
grade_this({
  if (!inherits(.result, "htest")) {
    fail("Run a t-test with the formula syntax: a number, a tilde, a group.")
  }
  if (isTRUE(all.equal(.result$p.value, .solution$p.value))) {
    pass("p = 0.13: no good evidence the arms differ in BMI. Learning to say 'no evidence of a difference' with a straight face is half of statistics.")
  }
  fail("Close. The formula is `bmi ~ arm`, with `data = patients`.")
})
```

**2.** Test whether smoking status (`smoker`) is associated with treatment `arm` in `patients`. The chi-squared test handles a 3-by-2 table without complaint.

```{r pr-chisq, exercise=TRUE}
# Your code here

```

```{r pr-chisq-hint-1}
# Two steps, one inside the other. Build the cross-tabulation of the two
# columns first with table(), then hand that to the test.
```

```{r pr-chisq-hint-2}
# chisq.test(table(patients$smoker, patients$arm))
```

```{r pr-chisq-solution}
chisq.test(table(patients$smoker, patients$arm))
```

```{r pr-chisq-check}
grade_this({
  if (!inherits(.result, "htest")) {
    fail("Build the counts with `table()`, then hand them to `chisq.test()`.")
  }
  if (isTRUE(all.equal(.result$p.value, .solution$p.value))) {
    pass("p = 0.53: smoking status falls evenly across the arms, as randomisation intended. Note the test took a 3-by-2 table in its stride.")
  }
  fail("Close. Table `smoker` against `arm` from patients, then run `chisq.test()` on it.")
})
```

**3.** Run the change-score t-test again (`change ~ arm` in `bp_change`) and pipe it into `tidy()`.

```{r pr-tidy, exercise=TRUE}
# Your code here

```

```{r pr-tidy-hint-1}
# Run the same change-score t-test as before, then pipe the whole result
# into the broom function that turns it into a data frame.
```

```{r pr-tidy-hint-2}
# t.test(change ~ arm, data = bp_change) |> tidy()
```

```{r pr-tidy-solution}
t.test(change ~ arm, data = bp_change) |>
  tidy()
```

```{r pr-tidy-check}
grade_this({
  pass_if_equal(message = "The trial's headline result, ready to report: a 4.8 mmHg bigger fall on the active arm, CI 0.2 to 9.4, p = 0.039.")
  if (inherits(.result, "htest")) {
    fail("The test itself is right. It just needs tidying: pipe it into `tidy()` to get the one-row data frame.")
  }
  fail("Not quite: run `t.test(change ~ arm, data = bp_change)` and pipe the result into `tidy()`.")
})
```

```{r pr-tidy-wrong, eval=FALSE, include=FALSE}
# expect: needs tidying
t.test(change ~ arm, data = bp_change)
```

## Recap

Two tests cover most clinical comparisons:

- **`t.test(number ~ group, data)`** compares a mean between two groups
- **`chisq.test(table(x, y))`** tests whether two categories are associated
- **`tidy()`** turns any result into a one-row data frame for reporting

And the discipline underneath: test the *outcome*, not the baseline, and report the confidence interval alongside the p-value. Next, regression: the tool for when one comparison at a time is not enough.

### Finished this lesson?

A lesson keeps hold of your R console while it is open, so closing the
browser tab is not enough to get it back. Press `Esc` with your cursor in
the **Console** (or click the red stop button at the top right of the
Console pane) to close the lesson down. Then:

```
learn()        # back to the menu
learn("m12")   # straight on to the next lesson
```

Nothing is lost by closing a lesson down and coming back to it later.
