---
title: "R for Clinical Statistics: Visualisation II"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Clinical plots that carry a message: trends over time, small multiples with faceting, and tidying a figure for a paper."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")

# Blood-pressure measurements over time, with each patient's arm attached.
bp <- labs |>
  filter(measure == "sbp") |>
  left_join(select(patients, id, arm), by = "id")

# The same, averaged to one value per arm per visit.
bp_mean <- bp |>
  group_by(arm, weeks) |>
  summarise(sbp = mean(value), .groups = "drop")
```

## Welcome

Lesson 7 covered the three plots you reach for most. This one turns them into figures that carry a clinical message: how a measure changes **over time**, how to compare arms with **small multiples**, and how to **tidy** a plot for a paper.

The promises hold: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away. We work with `bp`, the systolic blood-pressure readings at weeks 0, 12 and 24, with each patient's treatment arm attached.

## A first look over time

Press **Run Code** to plot every reading against the week it was taken:

```{r cloud, exercise=TRUE}
ggplot(bp, aes(x = weeks, y = value)) +
  geom_point()
```

A cloud of points. You can tell blood pressure drifts down, but you cannot follow any one patient, and the two arms are mixed together. Two moves fix that: connect each patient's own readings, and split by arm.

## Connecting each patient over time

To join up one patient's readings into a line, you tell ggplot which rows belong together with `group`, then draw with `geom_line()`. The grouping is set; **fill in the geom**:

```{r lines, exercise=TRUE, exercise.blanks="___+"}
ggplot(bp, aes(x = weeks, y = value, group = id)) +
  ____()
```

```{r lines-hint}
# The geom that draws a line through the points is:
# geom_line
```

```{r lines-solution}
ggplot(bp, aes(x = weeks, y = value, group = id)) +
  geom_line()
```

```{r lines-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Keep the `ggplot(...)` start and add a geom.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomLine" %in% geoms) {
    pass("One line per patient, thanks to `group = id`. Without that grouping ggplot would not know which points join up.")
  }
  fail("Almost: `geom_line()` draws the lines. The Hint has it.")
})
```

That tangle of lines has a name, a spaghetti plot, and it is honest: real patient data is noisy. To see the signal underneath, average it.

## The trend, one line per arm

`bp_mean` holds the mean blood pressure for each arm at each visit. **Fill in the column** to give each arm its own coloured line:

```{r trend, exercise=TRUE, exercise.blanks="___+"}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = ____)) +
  geom_line()
```

```{r trend-hint}
# You want one line per treatment arm, in its own colour.
# The column is called: arm
```

```{r trend-solution}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
  geom_line()
```

```{r trend-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if ("arm" %in% cols) {
    pass("There is the signal: both arms start together, and the active arm falls faster. That is the whole trial in one picture.")
  }
  fail("Almost: put `arm` in `colour = ` so each arm gets its own line.")
})
```

```{r trend-wrong, eval=FALSE, include=FALSE}
# expect: arm
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = weeks)) +
  geom_line()
```

## Small multiples with faceting

**Faceting** splits one crowded plot into a grid of small panels, one per group, every panel drawn to the same scale. That shared scale is the point: identical axes let you compare the panels directly by eye. The technique has a name of its own, *small multiples*.

Reach for it when a single panel is too busy to read. The spaghetti plot earlier is the case in point: a hundred lines on top of one another, where even colouring by arm would leave the two sets tangled together. Give each arm its own panel and half the crowding goes away. That is the trade in general. Colour keeps everything in one frame and works well while the groups stay easy to tell apart; faceting spends space to buy clarity, and wins as soon as the groups start hiding each other.

`facet_wrap()` does the splitting. Read `~ arm` as *"one panel per arm."* That squiggle is called a **tilde**, and it is R's way of saying "by": one panel *by* arm. You will meet it again in the analysis lessons, where models are written the same way.

**Fill in the column** to split the patient lines by arm:

```{r facet, exercise=TRUE, exercise.blanks="___+"}
ggplot(bp, aes(x = weeks, y = value, group = id)) +
  geom_line() +
  facet_wrap(~ ____)
```

```{r facet-hint}
# One panel per treatment arm, so facet by:
# arm
```

```{r facet-solution}
ggplot(bp, aes(x = weeks, y = value, group = id)) +
  geom_line() +
  facet_wrap(~ arm)
```

```{r facet-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  fvars <- names(c(.result$facet$params$facets,
                   .result$facet$params$rows,
                   .result$facet$params$cols))
  if ("arm" %in% fvars) {
    pass("Two panels, side by side, each arm on its own. Faceting is often clearer than colour when the lines would overlap.")
  }
  fail("Almost: `facet_wrap(~ arm)` splits the plot into one panel per arm.")
})
```

## Tidying for a paper

The plot is right, but the labels are raw column names and the grey background is busy. `labs()` sets readable titles, and a **theme** changes the overall look. The one below is `theme_minimal()`, which trades ggplot's grey panel for a light grid; it reproduces better in print, and it is the one to reach for first. Every theme comes as a layer you add with a `+`, like any other. **Run this** to see the same trend, tidied:

```{r polish, exercise=TRUE}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
  geom_line(linewidth = 1) +
  labs(
    title  = "Systolic blood pressure over time, by arm",
    x      = "Weeks since baseline",
    y      = "Mean systolic BP (mmHg)",
    colour = "Arm"
  ) +
  theme_minimal()
```

Same data, ready for a slide or a paper. The habit: get the plot right first, then add `labs()` and a theme last.

### Colours everyone can read

One more thing belongs in "ready for a paper", and almost nobody is taught it. Around one man in twelve has some form of red-green colour blindness, and ggplot's default colours do not allow for it.

With two groups the defaults hold up, which is why every plot so far has been fine. At three they stop holding up: the default red and green are close to indistinguishable to a reader with the commonest form, so two of your three lines become one. The plot still looks right to you.

The fix is a single layer. `scale_colour_brewer()` swaps in a palette built for categories:

```{r cvd-palette, exercise=TRUE}
ggplot(patients, aes(x = bmi, y = sbp, colour = smoker)) +
  geom_point() +
  scale_colour_brewer(palette = "Dark2") +
  labs(colour = "Smoking status") +
  theme_minimal()
```

Three points worth keeping:

- use **`scale_fill_brewer()`** instead when the colour fills a shape, as in bars and box plots
- for an **ordered** variable, or a continuous one, `scale_colour_viridis_d()` and `scale_colour_viridis_c()` do the same job
- neither needs an extra package; both come with `ggplot2`

Journals increasingly ask about this, and a figure nobody can read is a figure that failed at its one job.

## Practice

Three to finish, no blanks. Write each full plot yourself; **Hint** and **Solution** are there if you want them.

**1.** Draw the mean-blood-pressure trend from `bp_mean` (`weeks` on x, `sbp` on y), one line per `arm`.

```{r pr-trend, exercise=TRUE}
# Your code here

```

```{r pr-trend-hint-1}
# Use bp_mean, which already holds one value per arm per visit.
# Time goes across, the mean goes up, and mapping arm to colour gives
# you one line per arm without any extra work.
```

```{r pr-trend-hint-2}
# ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) + geom_line()
```

```{r pr-trend-solution}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
  geom_line()
```

```{r pr-trend-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start with `ggplot(bp_mean, aes(...))`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if (!("GeomLine" %in% geoms)) {
    fail("Use `geom_line()` to draw the trend.")
  }
  if (!("arm" %in% cols)) {
    fail("Map `arm` to `colour` so each arm has its own line.")
  }
  pass("The two-arm trend, written from scratch.")
})
```

**2.** Take the patient lines (`bp`, `weeks` vs `value`, grouped by `id`) and split them into one panel per `arm`.

```{r pr-facet, exercise=TRUE}
# Your code here

```

```{r pr-facet-hint-1}
# Start from the spaghetti plot: bp, with group = id so each patient gets
# their own line. Then add one more layer with a + to split it into panels.
```

```{r pr-facet-hint-2}
# ggplot(bp, aes(x = weeks, y = value, group = id)) + geom_line() + facet_wrap(~ arm)
```

```{r pr-facet-solution}
ggplot(bp, aes(x = weeks, y = value, group = id)) +
  geom_line() +
  facet_wrap(~ arm)
```

```{r pr-facet-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomLine" %in% geoms)) {
    fail("Draw the patient lines with `geom_line()` first.")
  }
  fvars <- names(c(.result$facet$params$facets,
                   .result$facet$params$rows,
                   .result$facet$params$cols))
  if (!("arm" %in% fvars)) {
    fail("Add `facet_wrap(~ arm)` to split into one panel per arm.")
  }
  pass("Patient trajectories, one panel per arm.")
})
```

**3.** Take the trend plot and add a y-axis label. Give `labs()` a `y` of your choosing.

```{r pr-labs, exercise=TRUE}
# Your code here

```

```{r pr-labs-hint-1}
# The plot is the trend one again. Add a labs() layer with a + and name
# the axis you want to rename, giving it the text a reader should see.
```

```{r pr-labs-hint-2}
# ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
#   geom_line() +
#   labs(y = "Mean systolic BP (mmHg)")
```

```{r pr-labs-solution}
ggplot(bp_mean, aes(x = weeks, y = sbp, colour = arm)) +
  geom_line() +
  labs(y = "Mean systolic BP (mmHg)")
```

```{r pr-labs-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  ylab <- .result$labels$y
  if (!is.null(ylab) && !(ylab %in% c("sbp", "value"))) {
    pass("A readable y-axis label. `labs()` is where you make a plot ready for someone else to read.")
  }
  fail("Almost: add `labs(y = \"...\")` with a label in your own words, not the raw column name.")
})
```

## Recap

Three moves that turn a plot into a figure:

- **`geom_line()` with `group`** follows something over time, one line per unit
- **`facet_wrap(~ variable)`** gives each group its own panel, the small-multiples trick
- **`labs()` and a theme** make it readable, added last once the plot is right

You have now seen the plot that shows a whole trial at a glance: two arms, starting together, parting over time. Next you turn those same variables into the summary table every paper opens with.

### Finished this lesson?

A lesson keeps hold of your R console while it is open, so closing the
browser tab is not enough to get it back. Press `Esc` with your cursor in
the **Console** (or click the red stop button at the top right of the
Console pane) to close the lesson down. Then:

```
learn()        # back to the menu
learn("m9")    # straight on to the next lesson
```

Nothing is lost by closing a lesson down and coming back to it later.
