---
title: "R for Clinical Statistics: Visualisation I"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Turn a table into a picture with ggplot2: the grammar of graphics, and your first scatter, box and bar plots."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
```

## Welcome

A table of numbers hides its story. A plot tells it at a glance. This lesson is your first with **ggplot2**, the tool most clinical R users reach for, and by the end you will have drawn three of the workhorse plots yourself.

The promises hold. Nothing here can break, red text is normal, and **Hint** then **Solution** are one click away. This lesson has an extra reward built in: every exercise ends with a picture appearing.

## The grammar of graphics

Every ggplot is built from three parts, always in the same order:

1. **the data**: which table to draw from, for example `patients`
2. **the aesthetics** (`aes`): which column maps to which part of the plot, for example `age` on the x-axis and `sbp` on the y-axis
3. **the geometry** (`geom`): how to draw it, as points, boxes, bars, and so on

You add the geometry with a `+`. Read the code below as *"take `patients`, put `age` across and `sbp` up, and draw it as points."* Press **Run Code**:

```{r first-plot, exercise=TRUE}
ggplot(patients, aes(x = age, y = sbp)) +
  geom_point()
```

There it is: one point per patient, age against systolic blood pressure. Change `geom_point()` to `geom_line()` and re-run if you are curious. The same three parts build every plot from here on.

## Scatter plots

A scatter plot shows how two numbers relate. Here we look at BMI against blood pressure. The data and aesthetics are set; **fill in the geom** that draws points:

```{r scatter, exercise=TRUE, exercise.blanks="___+"}
ggplot(patients, aes(x = bmi, y = sbp)) +
  ____()
```

```{r scatter-hint}
# The geom that draws one point per row is called:
# geom_point
```

```{r scatter-solution}
ggplot(patients, aes(x = bmi, y = sbp)) +
  geom_point()
```

```{r scatter-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Keep the `ggplot(...)` start and add a geom after the `+`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomPoint" %in% geoms) {
    pass("A scatter plot: one point per patient, BMI against blood pressure. The geom is the part that decides *how* the data is drawn.")
  }
  fail("Almost: `geom_point()` draws the points. The Hint has it if you need it.")
})
```

```{r scatter-wrong, eval=FALSE, include=FALSE}
# expect: geom_point
ggplot(patients, aes(x = bmi, y = sbp)) +
  geom_line()
```

## Showing a third variable with colour

Mapping a column to **colour** lets one plot carry a third variable. **Fill in the column** so the two treatment arms are coloured differently:

```{r colour, exercise=TRUE, exercise.blanks="___+"}
ggplot(patients, aes(x = bmi, y = sbp, colour = ____)) +
  geom_point()
```

```{r colour-hint}
# You want the two treatment arms in different colours.
# The column is called: arm
```

```{r colour-solution}
ggplot(patients, aes(x = bmi, y = sbp, colour = arm)) +
  geom_point()
```

```{r colour-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  cols <- unlist(lapply(maps, function(m) {
    if (is.null(m$colour)) NULL else rlang::as_label(m$colour)
  }))
  if ("arm" %in% cols) {
    pass("Each arm now has its own colour, and ggplot adds a legend for free. Mapping a column to `colour` is how you fit a third variable onto the plot.")
  }
  fail("Almost: put `arm` inside `colour = ` so the arms are coloured apart.")
})
```

### The key down the side

That strip on the right of the plot is the **legend**, and it is your colour key: ggplot built it automatically from the column you mapped, so you never have to tell the reader which colour means which arm.

It does take its title straight from the column name, which is rarely what you would want in a paper. `labs()` renames it. Press **Run Code**:

```{r colour-legend, exercise=TRUE}
ggplot(patients, aes(x = bmi, y = sbp, colour = arm)) +
  geom_point() +
  labs(colour = "Treatment arm")
```

`labs()` renames the axes the same way, and Lesson 8 puts it to work properly.

Anything inside `aes()` is *mapped from the data*. If instead you wanted every point the same fixed colour, you would set it outside `aes()`, like `geom_point(colour = "steelblue")`.

## Box plots

A box plot compares a number across groups. The box holds the middle half of patients, the line inside is the median. Here we compare blood pressure between the two arms. **Fill in the geom**:

```{r box, exercise=TRUE, exercise.blanks="___+"}
ggplot(patients, aes(x = arm, y = sbp)) +
  ____()
```

```{r box-hint}
# The geom that draws a box-and-whisker for each group is:
# geom_boxplot
```

```{r box-solution}
ggplot(patients, aes(x = arm, y = sbp)) +
  geom_boxplot()
```

```{r box-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomBoxplot" %in% geoms) {
    pass("One box per arm. The line is the median, the box the middle half of patients, so you can compare the groups at a glance.")
  }
  fail("Almost: `geom_boxplot()` draws a box for each group.")
})
```

## Bar charts

A bar chart counts how many fall into each category. Put a category on the x-axis and `geom_bar()` does the counting. **Fill in the geom** to count patients by smoking status:

```{r bar, exercise=TRUE, exercise.blanks="___+"}
ggplot(patients, aes(x = smoker)) +
  ____()
```

```{r bar-hint}
# The geom that draws a bar per category, counting the rows, is:
# geom_bar
```

```{r bar-solution}
ggplot(patients, aes(x = smoker)) +
  geom_bar()
```

```{r bar-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if ("GeomBar" %in% geoms) {
    pass("A bar per smoking group, its height the number of patients. `geom_bar()` counts the rows for you, so there is no y to set.")
  }
  fail("Almost: `geom_bar()` draws the bars and counts each group.")
})
```

## Practice

Three to finish, no blanks. Write each full plot yourself; **Hint** and **Solution** are there if you want them. Remember the three parts: data, `aes`, geom.

**1.** Draw a scatter plot of `age` (x-axis) against `bmi` (y-axis).

```{r pr-scatter, exercise=TRUE}
# Your code here

```

```{r pr-scatter-hint}
# ggplot(patients, aes(x = age, y = bmi)) + geom_point()
```

```{r pr-scatter-solution}
ggplot(patients, aes(x = age, y = bmi)) +
  geom_point()
```

```{r pr-scatter-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start with `ggplot(patients, aes(...))`.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomPoint" %in% geoms)) {
    fail("For a scatter plot, add `geom_point()`.")
  }
  pass("A scatter of age against BMI, written from scratch.")
})
```

**2.** Draw a box plot of `bmi` (y-axis) for each `smoker` group (x-axis).

```{r pr-box, exercise=TRUE}
# Your code here

```

```{r pr-box-hint}
# ggplot(patients, aes(x = smoker, y = bmi)) + geom_boxplot()
```

```{r pr-box-solution}
ggplot(patients, aes(x = smoker, y = bmi)) +
  geom_boxplot()
```

```{r pr-box-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomBoxplot" %in% geoms)) {
    fail("For a box plot, add `geom_boxplot()`.")
  }
  pass("BMI across the smoking groups, one box each.")
})
```

**3.** Draw a bar chart of the number of patients in each treatment `arm`.

```{r pr-bar, exercise=TRUE}
# Your code here

```

```{r pr-bar-hint}
# ggplot(patients, aes(x = arm)) + geom_bar()
```

```{r pr-bar-solution}
ggplot(patients, aes(x = arm)) +
  geom_bar()
```

```{r pr-bar-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot.")
  }
  geoms <- vapply(.result$layers, function(l) class(l$geom)[1], character(1))
  if (!("GeomBar" %in% geoms)) {
    fail("For a bar chart, add `geom_bar()`.")
  }
  pass("A bar per arm, each as tall as the group is large.")
})
```

## Recap

Every plot you drew today used the same three parts:

- **data**: the table, like `patients`
- **`aes()`**: which column goes where (x, y, colour)
- **`geom_`**: how to draw it, as points, boxes or bars

Swap the geom and you change the picture; swap the aesthetics and you change what it shows. That small grammar covers a great deal of ground. Next you shape these into publication-ready clinical figures.
