---
title: "R for Clinical Statistics: Summary Tables"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Build the 'Table 1' every clinical paper opens with: baseline characteristics by group, in one line, with finalfit."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(finalfit)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
```

## Welcome

Open almost any clinical paper and the first table describes who was in the study, split by group: mean age per arm, the sex balance, and so on. It is so standard it has a name, **Table 1**. This lesson builds one in a single line.

The promises hold: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away. We use the `patients` cohort and the `finalfit` package, built for exactly this job.

## A first Table 1

`summary_factorlist()` takes three things: the data, the **dependent** (the column to compare across, here the treatment `arm`), and the **explanatory** variables to describe. Press **Run Code**:

```{r first-table, exercise=TRUE}
patients |>
  summary_factorlist("arm", c("age", "sex"))
```

One row block per variable, one column per arm. Numbers like `age` come out as mean and standard deviation; categories like `sex` come out as counts and percentages. `finalfit` picks the right summary for each type automatically.

## Choosing what to compare across

The **dependent** is the column whose groups become the table's columns. **Fill it in** so the table compares the treatment arms, now across four variables:

```{r dependent, exercise=TRUE, exercise.blanks="___+"}
patients |>
  summary_factorlist("____", c("age", "sex", "smoker", "bmi"))
```

```{r dependent-hint}
# You want a column for each treatment arm, so compare across:
# "arm"
```

```{r dependent-solution}
patients |>
  summary_factorlist("arm", c("age", "sex", "smoker", "bmi"))
```

```{r dependent-check}
grade_this({
  pass_if_equal(message = "That is a full Table 1: every variable summarised for each arm, side by side, ready to drop into a paper.")
  fail("Not quite: the dependent is the column you compare across, written in quotes as `\"arm\"`.")
})
```

```{r dependent-wrong, eval=FALSE, include=FALSE}
# expect: arm
patients |>
  summary_factorlist("Arm", c("age", "sex", "smoker", "bmi"))
```

## Choosing what to describe

The **explanatory** variables are the list inside `c(...)`. Add to that list and the table grows a row block. **Fill in** the variable so smoking status joins the table:

```{r explanatory, exercise=TRUE, exercise.blanks="___+"}
patients |>
  summary_factorlist("arm", c("age", "sex", "____"))
```

```{r explanatory-hint}
# The column recording smoking status is called:
# smoker
```

```{r explanatory-solution}
patients |>
  summary_factorlist("arm", c("age", "sex", "smoker"))
```

```{r explanatory-check}
grade_this({
  pass_if_equal(message = "Smoking status now sits in the table, counted for each arm. Add as many variables to the list as you need.")
  fail("Not quite: add `\"smoker\"`, in quotes, to the list of variables to describe.")
})
```

## Adding a comparison

Sometimes you want a p-value for each row, testing whether the groups differ. Set `p = TRUE` and `finalfit` runs a sensible default test for each variable, a t-test for a number, a chi-squared for a category. Here we compare the men and women in the cohort. **Fill in** the value:

```{r pvalues, exercise=TRUE, exercise.blanks="___+"}
patients |>
  summary_factorlist("sex", c("age", "smoker", "bmi"), p = ____)
```

```{r pvalues-hint}
# You want the p-values switched on. The value is:
# TRUE
```

```{r pvalues-solution}
patients |>
  summary_factorlist("sex", c("age", "smoker", "bmi"), p = TRUE)
```

```{r pvalues-check}
grade_this({
  pass_if_equal(message = "A p-value column appears, one test per variable. Treat these as a guide: decide your comparisons in advance rather than reading down the column for anything below 0.05.")
  fail("Almost: switch the p-values on with `p = TRUE`, with no quotes.")
})
```

### One p-value per variable, not per level

Look at the `smoker` block in that table. Three levels, but only one p-value, sitting on the first row. That p belongs to the **whole variable**, not to the level it happens to line up with. It is a single chi-squared test asking one question: does the split across Never, Former and Current differ between the groups at all?

There is no option to get a p-value per level, and that is deliberate. Running three tests where one will do inflates your false-positive rate, and the answers would overlap anyway. When you do need to know how each level compares against a reference, the tool is regression: it gives you an estimate, a confidence interval and a p-value for every level measured against a baseline you choose. That is Lesson 12.

One argument worth knowing now: `p_cat = "fisher"` swaps the chi-squared for Fisher's exact test, which is the safer choice when any cell in the table has a small count.

### Why not p-values by arm?

You may have noticed we switched from `arm` to `sex` for that exercise. That was deliberate. In a randomised trial, any baseline difference between the arms is there by chance; the randomisation guarantees it. A significance test on the baseline table answers a question you already know the answer to, which is why the CONSORT guidance says to leave those p-values out. Describe the arms; test the outcomes.

For groups that were *not* assigned at random, like the men and women above, a baseline comparison can earn its place.

## Practice

Three to finish, no blanks. Write each table yourself; **Hint** and **Solution** are there if you want them.

**1.** Build a Table 1 of `age`, `bmi` and `smoker`, compared across the treatment `arm`.

```{r pr-table, exercise=TRUE}
# Your code here

```

```{r pr-table-hint}
# patients |> summary_factorlist("arm", c("age", "bmi", "smoker"))
```

```{r pr-table-solution}
patients |>
  summary_factorlist("arm", c("age", "bmi", "smoker"))
```

```{r pr-table-check}
grade_this({
  if (!is.data.frame(.result) || !"label" %in% names(.result)) {
    fail("That hasn't produced a summary table. Start from `patients |> summary_factorlist(...)`.")
  }
  pass_if_equal(message = "A tidy three-variable Table 1, written from scratch.")
  vars <- setdiff(unique(.result$label), "")
  if (!"Active" %in% names(.result)) {
    fail("The columns should be the two treatment arms, so the dependent is `\"arm\"`.")
  }
  if (!setequal(vars, c("age", "bmi", "smoker"))) {
    fail("Check the variable list: this one wants `age`, `bmi` and `smoker` inside `c(...)`, each in quotes.")
  }
  if ("p" %in% names(.result)) {
    fail("The table is right. This exercise doesn't ask for p-values, though, so drop `p = TRUE`.")
  }
  fail("Not quite. Aim for `age`, `bmi` and `smoker` described across the treatment `arm`.")
})
```

```{r pr-table-wrong, eval=FALSE, include=FALSE}
# expect: dependent is
patients |>
  summary_factorlist("sex", c("age", "bmi", "smoker"))
```

**2.** Now compare those same three variables across `sex` instead of arm.

```{r pr-bysex, exercise=TRUE}
# Your code here

```

```{r pr-bysex-hint}
# Swap the dependent from "arm" to "sex":
# patients |> summary_factorlist("sex", c("age", "bmi", "smoker"))
```

```{r pr-bysex-solution}
patients |>
  summary_factorlist("sex", c("age", "bmi", "smoker"))
```

```{r pr-bysex-check}
grade_this({
  if (!is.data.frame(.result) || !"label" %in% names(.result)) {
    fail("That hasn't produced a summary table. Start from `patients |> summary_factorlist(...)`.")
  }
  pass_if_equal(message = "Same variables, split by sex now. Changing the dependent is all it takes to re-point the table.")
  vars <- setdiff(unique(.result$label), "")
  if (!"Female" %in% names(.result)) {
    fail("The columns should be Female and Male, so the dependent is `\"sex\"`.")
  }
  if (!setequal(vars, c("age", "bmi", "smoker"))) {
    fail("Keep the same three variables as the last exercise: `age`, `bmi` and `smoker`.")
  }
  if ("p" %in% names(.result)) {
    fail("So close: the table is exactly right, but this one doesn't ask for a p-value column. Drop `p = TRUE` and you have it. You switch them back on in the next exercise.")
  }
  fail("Not quite. Aim for `age`, `bmi` and `smoker` described across `sex`.")
})
```

```{r pr-bysex-wrong, eval=FALSE, include=FALSE}
# expect: drop `p = TRUE`
patients |>
  summary_factorlist("sex", c("age", "bmi", "smoker"), p = TRUE)
```

**3.** Build a table of `age` and `bmi` compared across `sex`, with p-values switched on.

```{r pr-pvalues, exercise=TRUE}
# Your code here

```

```{r pr-pvalues-hint}
# patients |> summary_factorlist("sex", c("age", "bmi"), p = TRUE)
```

```{r pr-pvalues-solution}
patients |>
  summary_factorlist("sex", c("age", "bmi"), p = TRUE)
```

```{r pr-pvalues-check}
grade_this({
  if (!is.data.frame(.result) || !"label" %in% names(.result)) {
    fail("That hasn't produced a summary table. Start from `patients |> summary_factorlist(...)`.")
  }
  pass_if_equal(message = "Two variables compared between the sexes, each with a p-value. And you know when to leave that column out: a randomised baseline table describes, it does not test.")
  vars <- setdiff(unique(.result$label), "")
  if (!"Female" %in% names(.result)) {
    fail("The columns should be Female and Male, so the dependent is `\"sex\"`.")
  }
  if (!setequal(vars, c("age", "bmi"))) {
    fail("This one wants just two variables: `age` and `bmi`.")
  }
  if (!"p" %in% names(.result)) {
    fail("The table is right, but there's no p-value column yet. Add `p = TRUE`.")
  }
  fail("Not quite. Aim for `age` and `bmi` compared across `sex`, with `p = TRUE`.")
})
```

```{r pr-pvalues-wrong, eval=FALSE, include=FALSE}
# expect: no p-value column
patients |>
  summary_factorlist("sex", c("age", "bmi"))
```

## Recap

The whole of Table 1 in one function:

- **`summary_factorlist(dependent, explanatory)`** describes each explanatory variable for every group of the dependent
- **numbers** become mean and SD, **categories** become counts and percentages, chosen for you
- **`p = TRUE`** adds a comparison test per row, for groups that were not assigned at random; a randomised baseline table carries none (CONSORT)

`finalfit` is built for clinical reporting, and this is its everyday job. Next, Part 3: a closer look at describing data itself — which summary to use, and when the mean misleads.
