---
title: "Systolic blood pressure trial: analysis report"
author: "Your name here"
date: today
format: html
---

<!-- This is your report to finish. Each TODO chunk below matches an exercise
     you completed in Lesson 14: paste your code in, press Render, and watch
     the section appear. The document renders at every stage, so render early
     and render often. Stuck? r4clinstats::start_report(completed = TRUE)
     gives you the finished version to compare against. -->

```{r}
#| label: setup
#| message: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(survival)
library(finalfit)
library(knitr)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")
data("outcomes", package = "r4clinstats")

# Change from baseline to week 24, one row per patient, with Placebo as the
# reference level so model coefficients read as treatment effects.
bp_change <- labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value, names_prefix = "v") |>
  left_join(select(patients, id, arm), by = "id") |>
  mutate(
    change = v2 - v0,
    arm = relevel(factor(arm), ref = "Placebo")
  ) |>
  select(id, arm, baseline = v0, week24 = v2, change)

# Mean blood pressure per arm per visit, for the trend figure.
bp_mean <- labs |>
  filter(measure == "sbp") |>
  left_join(select(patients, id, arm), by = "id") |>
  group_by(arm, weeks) |>
  summarise(sbp = mean(value), .groups = "drop")

# Outcomes joined to patient details.
trial <- outcomes |>
  left_join(select(patients, id, age, sex, arm), by = "id") |>
  mutate(
    arm = relevel(factor(arm), ref = "Placebo"),
    response = factor(response)
  )
```

## Background

This report analyses a randomised, two-arm trial of an antihypertensive drug against placebo: systolic blood pressure over 24 weeks, treatment response, and event-free survival over one year. The data are synthetic teaching data from the `r4clinstats` course and contain no real patients.

Everything in this document, every number, table and figure, is computed from the data at the moment the report is rendered. Change the data or the analysis and re-render, and the whole report rebuilds itself.

## Patients

The trial randomised `r nrow(patients)` patients, `r sum(patients$arm == "Active")` to the active arm and `r sum(patients$arm == "Placebo")` to placebo. Baseline characteristics were balanced across the arms, as randomisation intends.

```{r}
#| label: table-one
# TODO (Lesson 14, "The baseline table"): build the Table 1 comparing
# age, sex, smoker and bmi across the arms, then pipe it into kable().
```

## Blood pressure over 24 weeks

Mean systolic blood pressure fell in both arms, and fell faster on the drug.

```{r}
#| label: bp-trend
# TODO (Lesson 14, "The trend figure"): the mean blood pressure trend from
# bp_mean, one coloured line per arm, with readable labs() and a theme.
```

The primary analysis models each patient's change from baseline, adjusted for their baseline value.

```{r}
#| label: bp-model
# TODO (Lesson 14, "The primary analysis"): fit the baseline-adjusted model
# of change, tidy it with a confidence interval, and pipe it into kable().
```

TODO: once the model chunk above works, replace this sentence with the result
in words, using inline code for the numbers.

## Treatment response

`r sum(trial$response == "Y" & trial$arm == "Active")` of `r sum(trial$arm == "Active")` patients responded on the active arm, against `r sum(trial$response == "Y" & trial$arm == "Placebo")` of `r sum(trial$arm == "Placebo")` on placebo.

```{r}
#| label: response-model
# TODO (Lesson 14, "The odds ratio"): the logistic regression of response on
# arm, tidied into an odds ratio with confidence interval, piped into kable().
```

## Survival

```{r}
#| label: km-curve
# TODO (Lesson 14, "The survival curve"): fit the Kaplan-Meier estimate by
# arm, then draw the staircases with tidy() + ggplot + geom_step().
```

```{r}
#| label: cox-model
# TODO (Lesson 14, "The hazard ratio"): the Cox model of the treatment
# effect, tidied into a hazard ratio with confidence interval, via kable().
```

## Conclusion

TODO: two or three sentences in your own words. What did the drug do to blood
pressure, to response, and to survival? Use inline code for any number you
quote, so the conclusion can never drift out of step with the analysis.
