---
title: "R for Clinical Statistics: Tidy Data in Practice"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Reshape between long and wide with pivot, and join two tables: the moves that make repeated measurements easy."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(tidyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")
```

## Welcome

This is where tidy data earns its keep. You'll meet a second table, `labs`, that records the same patients measured again and again over time, and you'll learn the two moves that make that kind of data easy: **pivoting** (reshaping between long and wide) and **joining** (linking two tables on a shared column).

The usual promises hold: nothing here can break, red text is normal, and **Hint** then **Solution** are a click away.

## Meet the labs

Press **Run Code** to see the new table:

```{r meet-labs, exercise=TRUE}
glimpse(labs)
```

This is **long** data: one row per measurement, not one row per patient. Each patient appears several times, once for each `visit` (0, 1 and 2) and each `measure` (systolic blood pressure and weight). It is perfectly tidy, and yet this shape makes some questions awkward, as you're about to see.

## The shape problem

Say you want each patient's change in blood pressure from their first visit to their last. In the long shape, visit 0 and visit 2 sit on *different rows*, so there is no single row to subtract one from the other. Take a look at just the blood-pressure rows:

```{r filter-sbp, exercise=TRUE}
labs |>
  filter(measure == "sbp")
```

Three rows per patient, stacked. To compare visits, you want them spread across one row each. That is a job for `pivot_wider()`.

## Reshaping with `pivot_wider()`

`pivot_wider()` turns rows into columns, and it needs two things from you: which column supplies the new **headings**, and which supplies the **numbers** that fill them. We first narrow to the three columns we care about (`id`, `visit`, `value`), then spread the visits out. **Fill in both blanks**:

```{r pivot-wide, exercise=TRUE, exercise.blanks="___+"}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = ____, values_from = ____)
```

```{r pivot-wide-hint}
# You want one new column per visit, so the headings come from: visit
# The blood-pressure numbers that fill them live in: value
```

```{r pivot-wide-solution}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value)
```

```{r pivot-wide-check}
grade_this({
  pass_if_equal(message = "That's the payoff: one row per patient, a column for each visit. `names_from` picks the column whose values become headings, `values_from` picks what fills the cells. Visits 0 and 2 now sit side by side, ready to subtract.")
  if (is.data.frame(.result) && ncol(.result) > 10) {
    fail("Those two are the wrong way round. The headings come from `visit` (three of them), and the numbers that fill the cells come from `value`.")
  }
  fail("Not quite: one new column per visit, so `names_from = visit`, and the numbers to spread are in `value`.")
})
```

```{r pivot-wide-wrong, eval=FALSE, include=FALSE}
# expect: wrong way round
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = value, values_from = visit)
```

## Linking two tables with `left_join()`

`labs` holds the measurements but little about the patients themselves. Their age, sex and arm live over in `patients`. A **join** brings the two together, matching rows by a column they share, here `id`. Run this:

```{r join-demo, exercise=TRUE}
labs |>
  left_join(patients, by = "id")
```

Every lab row now carries that patient's `age`, `sex`, `arm` and the rest, looked up from `patients`. `left_join()` keeps every row on the left (`labs`) and adds the matching columns from the right.

Your turn, and this is where the join earns its keep. Once the tables are linked you can filter on a column that was never in `labs` to begin with. **Fill in the column** that keeps only the measurements from the Active arm:

```{r join-filter, exercise=TRUE, exercise.blanks="___+"}
labs |>
  left_join(patients, by = "id") |>
  filter(____ == "Active")
```

```{r join-filter-hint}
# The treatment arm arrived from `patients` in the join. The column is called: arm
```

```{r join-filter-solution}
labs |>
  left_join(patients, by = "id") |>
  filter(arm == "Active")
```

```{r join-filter-check}
grade_this({
  pass_if_equal(message = "{nrow(.result)} measurements, all from the Active arm. Note what happened there: you filtered on `arm`, a column `labs` never had. That is the point of joining, and every verb from the last two lessons works on the result.")
  if (is.data.frame(.result) && nrow(.result) == 0) {
    fail("That came back empty, which usually means the text didn't match. R is case-sensitive, so the arm is `\"Active\"`.")
  }
  fail("Not quite: the treatment arm came across from `patients` in the join, in a column called `arm`.")
})
```

```{r join-filter-wrong, eval=FALSE, include=FALSE}
# expect: case-sensitive
labs |>
  left_join(patients, by = "id") |>
  filter(arm == "active")
```

## When a join finds nothing

A join only works if the key matches on both sides, and it can go wrong in two very different ways.

If the two tables *name* the key differently, one `id` and the other `patient_id`, R stops and tells you: `Join columns in 'x' must be present in the data`. Loud, and easy to fix.

The other way is quiet, and far more dangerous. If the names match but the *values* don't, the join succeeds and returns nothing useful. Here the ids have picked up a stray trailing space, the kind of thing a spreadsheet does without telling you. Press **Run Code**:

```{r join-broken, exercise=TRUE}
labs_messy <- labs |>
  mutate(id = paste0(id, " "))     # "P001" is now "P001 "

labs_messy |>
  left_join(patients, by = "id") |>
  select(id, visit, measure, age, sex, arm)
```

Every row is still there, no error, no warning. But `age`, `sex` and `arm` are `NA` all the way down, because `"P001 "` never matched `"P001"`. A join that silently returns `NA` is easy to miss and easy to analyse by accident, so check the result after any join: if a column you expected is entirely `NA`, the key is the first suspect.

The fix is the design habit from earlier: give a column the same name and the same type in every table you will join on, and settle it once when you set the tables up.

## Practice

Two tables and the two new moves. Write each one yourself; the **Hint** and **Solution** are there when you want them.

**1.** Earlier you spread the blood-pressure rows. Do the same for **weight**: keep the weight rows, narrow to `id`, `visit` and `value`, then spread the visits across columns.

```{r pr-pivot, exercise=TRUE}
# Your code here

```

```{r pr-pivot-hint}
# Same shape as the sbp example, with measure == "weight":
# labs |> filter(measure == "weight") |> select(id, visit, value) |>
#   pivot_wider(names_from = visit, values_from = value)
```

```{r pr-pivot-solution}
labs |>
  filter(measure == "weight") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value)
```

```{r pr-pivot-check}
grade_this({
  pass_if_equal(message = "Same move as the blood-pressure table, now for weight: one row per patient, a column per visit.")
  if (is.data.frame(.result)) {
    if ("measure" %in% names(.result)) {
      fail("Nearly. Narrow to just `id`, `visit` and `value` with `select()` before pivoting, or the extra columns split the rows up.")
    }
    if (!all(c("0", "1", "2") %in% names(.result))) {
      fail("The visits should become the column headings, so `names_from = visit`.")
    }
  }
  fail("Not quite: keep the weight rows, narrow to `id`, `visit`, `value`, then `pivot_wider(names_from = visit, values_from = value)`.")
})
```

```{r pr-pivot-wrong, eval=FALSE, include=FALSE}
# expect: select
labs |>
  filter(measure == "weight") |>
  pivot_wider(names_from = visit, values_from = value)
```

**2.** Add each patient's treatment `arm` onto the `labs` table. Keep every lab row, and bring across only the `arm` column.

```{r pr-join, exercise=TRUE}
# Your code here

```

```{r pr-join-hint}
# Select just the columns you need from patients before joining:
# labs |> left_join(select(patients, id, arm), by = "id")
```

```{r pr-join-solution}
labs |>
  left_join(select(patients, id, arm), by = "id")
```

```{r pr-join-check}
grade_this({
  pass_if_equal(message = "Each lab row now carries its patient's arm. Selecting `id` and `arm` first keeps the join tidy.")
  if (is.data.frame(.result)) {
    if (!"arm" %in% names(.result)) {
      fail("The `arm` column hasn't come across yet. `left_join()` pulls it from `patients`, matching on `\"id\"`.")
    }
    if (length(setdiff(names(.result), c(names(labs), "arm")))) {
      fail("The join worked, but it brought every patient column with it. Narrow the right-hand table first: `left_join(select(patients, id, arm), by = \"id\")`.")
    }
  }
  fail("Not quite: join `select(patients, id, arm)` onto `labs` by `\"id\"`, so only the arm comes across.")
})
```

```{r pr-join-wrong, eval=FALSE, include=FALSE}
# expect: every patient column
labs |>
  left_join(patients, by = "id")
```

**3.** Using a join, count how many lab measurements fall under each treatment `arm`.

```{r pr-count-arm, exercise=TRUE}
# Your code here

```

```{r pr-count-arm-hint}
# Join first, then count. You can use the full patients table here:
# labs |> left_join(patients, by = "id") |> count(arm)
```

```{r pr-count-arm-solution}
labs |>
  left_join(patients, by = "id") |>
  count(arm)
```

```{r pr-count-arm-check}
grade_this({
  pass_if_equal(message = "Join then count: the measurements split by treatment arm. Two tables, one tally.")
  if (is.data.frame(.result) && setequal(names(.result), c("arm", "n")) &&
      sum(.result$n) != nrow(labs)) {
    fail("That has counted patients rather than measurements. Start from `labs` so every lab row is kept, join `patients` on, then `count(arm)`.")
  }
  fail("Not quite: join `patients` onto `labs` by `\"id\"`, then `count(arm)`.")
})
```

```{r pr-count-arm-wrong, eval=FALSE, include=FALSE}
# expect: counted patients
patients |>
  count(arm)
```

## Recap

Two moves, both about *shape* rather than content:

- **`pivot_wider()`** spreads one column into many, and **`pivot_longer()`** does the reverse
- **`left_join()`** links two tables on a shared column

Reshaping and joining are what make repeated-measures and multi-table clinical data workable, and both go far more smoothly when the tables were tidy to begin with. That thread runs through everything so far.
