---
title: "R for Clinical Statistics: Tidy Data in Practice"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Reshape between long and wide with pivot, and join two tables: the moves that make repeated measurements easy."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(tidyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("labs", package = "r4clinstats")
```

## Welcome

This is where tidy data earns its keep. You'll meet a second table, `labs`, that records the same patients measured again and again over time, and you'll learn the two moves that make that kind of data easy: **pivoting** (reshaping between long and wide) and **joining** (linking two tables on a shared column).

The usual promises hold: nothing here can break, red text is normal, and **Hint** then **Solution** are a click away.

## Meet the labs

Press **Run Code** to see the new table:

```{r meet-labs, exercise=TRUE}
glimpse(labs)
```

This is **long** data: one row per measurement, not one row per patient. Each patient appears several times, once for each `visit` (0, 1 and 2) and each `measure` (systolic blood pressure and weight). It is perfectly tidy, and yet this shape makes some questions awkward, as you're about to see.

## The shape problem

Say you want each patient's change in blood pressure from their first visit to their last. In the long shape, visit 0 and visit 2 sit on *different rows*, so there is no single row to subtract one from the other. Take a look at just the blood-pressure rows:

```{r filter-sbp, exercise=TRUE}
labs |>
  filter(measure == "sbp")
```

Three rows per patient, stacked. To compare visits, you want them spread across one row each. That is a job for `pivot_wider()`.

## Reshaping with `pivot_wider()`

`pivot_wider()` turns rows into columns. We first narrow to the three columns we care about (`id`, `visit`, `value`), then spread the visits out. **Fill in which column holds the numbers** to spread:

```{r pivot-value, exercise=TRUE, exercise.blanks="___+"}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = ____)
```

```{r pivot-value-hint}
# The blood-pressure numbers to spread across the new columns live in: value
```

```{r pivot-value-solution}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value)
```

```{r pivot-value-check}
grade_this({
  pass_if_equal(message = "That's the payoff: one row per patient, a column for each visit. Visits 0 and 2 now sit side by side, ready to subtract.")
  fail("Almost: the numbers to spread out are in the `value` column.")
})
```

Two blanks live in that call. **Now fill in which column supplies the new column names**, one per visit:

```{r pivot-names, exercise=TRUE, exercise.blanks="___+"}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = ____, values_from = value)
```

```{r pivot-names-hint}
# You want one new column per visit, so the names come from: visit
```

```{r pivot-names-solution}
labs |>
  filter(measure == "sbp") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value)
```

```{r pivot-names-check}
grade_this({
  pass_if_equal(message = "Exactly. `names_from` picks the column whose values become new headings; `values_from` picks what fills the cells.")
  fail("Not quite: you want one column per visit, so `names_from` is `visit`.")
})
```

## Linking two tables with `left_join()`

`labs` holds the measurements but little about the patients themselves. Their age, sex and arm live over in `patients`. A **join** brings the two together, matching rows by a column they share, here `id`. Run this:

```{r join-demo, exercise=TRUE}
labs |>
  left_join(patients, by = "id")
```

Every lab row now carries that patient's `age`, `sex`, `arm` and the rest, looked up from `patients`. `left_join()` keeps every row on the left (`labs`) and adds the matching columns from the right.

Your turn. **Fill in the column the two tables share** so each lab row picks up its patient details:

```{r join-key, exercise=TRUE, exercise.blanks="___+"}
labs |>
  left_join(patients, by = "____")
```

```{r join-key-hint}
# Both tables carry a patient identifier column called: id
```

```{r join-key-solution}
labs |>
  left_join(patients, by = "id")
```

```{r join-key-check}
grade_this({
  pass_if_equal(message = "Joined. `id` is the link the tables have in common, so R lines them up on it.")
  fail("Almost: the column both tables share is `id`.")
})
```

## When a join finds nothing

A join only works if the key matches on both sides. If one table calls the column `id` (`P001`, `P002`, and so on) while the other calls it `patient_id`, or stores it as a plain number, R finds no matches and every added column comes back empty (`NA`). The fix is the design habit from earlier: give a column the same name and the same type in every table you will join on, and settle it once when you set the tables up.

## Practice

Two tables and the two new moves. Write each one yourself; the **Hint** and **Solution** are there when you want them.

**1.** Earlier you spread the blood-pressure rows. Do the same for **weight**: keep the weight rows, narrow to `id`, `visit` and `value`, then spread the visits across columns.

```{r pr-pivot, exercise=TRUE}
# Your code here

```

```{r pr-pivot-hint}
# Same shape as the sbp example, with measure == "weight":
# labs |> filter(measure == "weight") |> select(id, visit, value) |>
#   pivot_wider(names_from = visit, values_from = value)
```

```{r pr-pivot-solution}
labs |>
  filter(measure == "weight") |>
  select(id, visit, value) |>
  pivot_wider(names_from = visit, values_from = value)
```

```{r pr-pivot-check}
grade_this({
  pass_if_equal(message = "Same move as the blood-pressure table, now for weight: one row per patient, a column per visit.")
  fail("Close. Keep the weight rows, narrow to `id`, `visit`, `value`, then `pivot_wider(names_from = visit, values_from = value)`.")
})
```

**2.** Add each patient's treatment `arm` onto the `labs` table. Keep every lab row, and bring across only the `arm` column.

```{r pr-join, exercise=TRUE}
# Your code here

```

```{r pr-join-hint}
# Select just the columns you need from patients before joining:
# labs |> left_join(select(patients, id, arm), by = "id")
```

```{r pr-join-solution}
labs |>
  left_join(select(patients, id, arm), by = "id")
```

```{r pr-join-check}
grade_this({
  pass_if_equal(message = "Each lab row now carries its patient's arm. Selecting `id` and `arm` first keeps the join tidy.")
  fail("Almost. Join `select(patients, id, arm)` onto `labs` by `\"id\"`, so only the arm comes across.")
})
```

**3.** Using a join, count how many lab measurements fall under each treatment `arm`.

```{r pr-count-arm, exercise=TRUE}
# Your code here

```

```{r pr-count-arm-hint}
# Join first, then count. You can use the full patients table here:
# labs |> left_join(patients, by = "id") |> count(arm)
```

```{r pr-count-arm-solution}
labs |>
  left_join(patients, by = "id") |>
  count(arm)
```

```{r pr-count-arm-check}
grade_this({
  pass_if_equal(message = "Join then count: the measurements split by treatment arm. Two tables, one tally.")
  fail("Close. Join `patients` onto `labs` by `\"id\"`, then `count(arm)`.")
})
```

## Recap

Two moves, both about *shape* rather than content:

- **`pivot_wider()`** spreads one column into many, and **`pivot_longer()`** does the reverse
- **`left_join()`** links two tables on a shared column

Reshaping and joining are what make repeated-measures and multi-table clinical data workable, and both go far more smoothly when the tables were tidy to begin with. That thread runs through everything so far.
