---
title: "Crib sheet: every function in the course"
---

```{r crib-helpers, include=FALSE}
# Each entry is written as a template with the learner's parts in braces, for
# example "glimpse({your_data})". The braces become highlighted, italic words
# and everything else is shown exactly as it is typed.
#
# The tables are built as HTML rather than markdown for two reasons: markdown
# cannot highlight part of a piece of code, and it cannot show a pipe (|>)
# inside code in a table cell at all.
esc <- function(x) htmltools::htmlEscape(x)

yours <- function(x) gsub("\\{([^}]+)\\}", "<mark>\\1</mark>", x)

# Descriptions allow `code`, **bold** and {placeholders}, nothing more.
md <- function(x) {
  x <- gsub("`([^`]+)`", "<code>\\1</code>", esc(x))
  gsub("\\*\\*([^*]+)\\*\\*", "<strong>\\1</strong>", yours(x))
}

# One entry: the template, what it does, an example from the lessons (left
# out when the template has nothing to fill in), and the lesson it came from.
entry <- function(code, what, eg = NULL, lesson = "") {
  list(code = code, what = what, eg = eg, lesson = lesson)
}

crib <- function(..., lessons = TRUE) {
  cells <- vapply(list(...), function(e) {
    eg <- if (is.null(e$eg)) "" else
      sprintf('<div class="crib-eg">e.g. <code>%s</code></div>', esc(e$eg))
    last <- if (lessons) sprintf("<td>%s</td>", e$lesson) else ""
    sprintf("<tr><td><code>%s</code>%s</td><td>%s</td>%s</tr>",
            yours(esc(e$code)), eg, md(e$what), last)
  }, character(1))
  head <- paste0("<th>Code</th><th>What it does</th>",
                 if (lessons) "<th>Lesson</th>" else "")
  htmltools::HTML(paste0(
    '<div class="table-responsive"><table class="table crib">',
    "<thead><tr>", head, "</tr></thead>",
    "<tbody>", paste(cells, collapse = ""), "</tbody></table></div>"
  ))
}
```

```{=html}
<style>
.crib mark, .crib-key mark {
  background: #fff3a3;
  color: inherit;
  font-style: italic;
  padding: 0 0.15em;
  border-radius: 2px;
}
.crib td { vertical-align: top; }
.crib td:first-child { width: 45%; }
.crib td:first-child code { white-space: normal; }
.crib-eg { margin-top: 0.35em; font-size: 0.85em; color: #6c757d; }
.crib-eg code { color: inherit; background: none; padding: 0; }
@media print {
  .crib mark, .crib-key mark {
    background: #fff3a3 !important;
    -webkit-print-color-adjust: exact;
    print-color-adjust: exact;
  }
}
</style>
```

Every function the course teaches, on one page, grouped by the job you want done.

Keep this open in a second browser tab while you work. A lesson takes over your R console while it runs, so you cannot look things up there, but this page is always to hand. It is also worth printing.

## How to read an entry

<p class="crib-key">Each line of code is written so it works on any data, not just the course's. The <mark>highlighted</mark> words are yours: replace each one with a name from your own data, such as your table or one of its columns. Everything else, including the brackets, commas and quote marks, is typed exactly as shown. Where a line has nothing highlighted, type it as it stands.</p>

Under each one, in grey, is an example from the lessons, and the last column says which lesson taught it, so you can go back to the full explanation whenever a one-line reminder is not enough.

## Two symbols worth knowing

These two turn up everywhere and neither is guessable.

**The pipe, `|>`, means "and then."** It carries the result of one step into the next, so a pipeline reads top to bottom like a recipe:

```r
patients |>
  filter(age > 70) |>
  select(id, age, arm)
```

*Take patients, and then keep the older ones, and then show these three columns.*

**The tilde, `~`, means "by", or "explained by."** What you are measuring goes on the left, what splits or explains it goes on the right:

```r
change ~ arm          # change, by arm
facet_wrap(~ arm)     # one panel, by arm
```

A third piece of punctuation is worth knowing even though it is not a symbol you type often. **`package::function()`**, with two colons, means "this function, out of this package":

```r
r4clinstats::start_report()        # start_report(), from r4clinstats
dplyr::filter(patients, age > 65)  # filter(), from dplyr
```

It is a command to run in the Console. It is not a web address, and there is nothing to click.

## Getting your data in

```{r, echo=FALSE}
crib(
  entry('read_csv(here("{folder}", "{file_name.csv}"))',
        "Reads a CSV. `here()` builds the path from the top of your project, so it works on any computer. Each part of the path is its own quoted name.",
        'read_csv(here("data", "linelist.csv"))', "M6"),
  entry('read_excel(here("{folder}", "{file_name.xlsx}"))',
        "Reads a spreadsheet directly, no export step. From the `readxl` package.",
        'read_excel(here("data", "trial-data.xlsx"))', "M6"),
  entry('read_excel({path}, sheet = "{sheet_name}")',
        "Picks one sheet out of a workbook. The {path} is the whole `here(...)` from the line above.",
        'read_excel(here("data", "trial.xlsx"), sheet = "Baseline")', "M6"),
  entry("read_excel({path}, skip = {number})",
        "Ignores that many rows above the real headers, such as a title or logo.",
        'read_excel(here("data", "trial.xlsx"), skip = 2)', "M6"),
  entry('read_excel({path}, na = c("{code}", "{code}"))',
        "Says what should count as missing. List every code your spreadsheet uses, each in its own quotes. `\"\"` is an empty cell.",
        'read_excel(here("data", "trial.xlsx"), na = c("", "N/A", "-"))', "M6")
)
```

Read the file, never edit it. Then `glimpse()` the result and check the row count and column types before going further.

## Looking at your data

Start here when you meet a new table.

```{r, echo=FALSE}
crib(
  entry("glimpse({your_data})",
        "Every column, its type, and the first few values. The best first look.",
        "glimpse(patients)", "M2"),
  entry("head({your_data})",
        "The first six rows.",
        "head(patients)", "M2"),
  entry("summary({your_data}${column})",
        "Min, max, quartiles, median and mean of a numeric column. The `$` picks one column out of a table.",
        "summary(patients$age)", "M10"),
  entry("class({your_data}${column})",
        "What type a column is: `chr`, `dbl`, `fct`, `lgl`.",
        "class(patients$sex)", "M2"),
  entry("table({your_data}${column})",
        "Counts of each category, base R's version of `count()`.",
        "table(patients$smoker)", "M2"),
  entry("nrow({your_data})",
        "How many rows.",
        "nrow(patients)", "M14")
)
```

## Choosing rows and columns

```{r, echo=FALSE}
crib(
  entry('filter({your_data}, {column} == "{value}")',
        "Keep the rows that match. Note the double `==`. The quote marks are there because the value is text; for a number, leave them off: `filter({your_data}, {column} > {number})`.",
        'filter(patients, arm == "Active")', "M3"),
  entry("filter({your_data}, {condition}, {condition})",
        "Several conditions at once, separated by commas. A row is kept only if every one holds. Each condition is written like the ones above.",
        'filter(patients, age > 70, sex == "Female")', "M6"),
  entry("select({your_data}, {column}, {column})",
        "Keep only these columns, in this order. List as many as you need.",
        "select(patients, id, age, arm)", "M3"),
  entry("arrange({your_data}, {column})",
        "Sort, smallest first.",
        "arrange(patients, age)", "M3"),
  entry("arrange({your_data}, desc({column}))",
        "Sort, largest first. `desc()` flips it.",
        "arrange(patients, desc(age))", "M3")
)
```

## Counting and summarising

```{r, echo=FALSE}
crib(
  entry("count({your_data}, {column})",
        "How many rows in each group. The count column is called `n`.",
        "count(patients, smoker)", "M3"),
  entry("summarise({your_data}, {new_name} = mean({column}))",
        "Collapse a column to one number. You choose the {new_name}, and it becomes the column's heading.",
        "summarise(patients, mean_age = mean(age))", "M4"),
  entry("mean({column})",
        "The summary numbers themselves: `median()`, `sd()` and `IQR()` are written the same way. Use mean and SD for a symmetric shape, median and IQR when it is lopsided.",
        "mean(age)", "M4, M10")
)
```

The two you will reach for most are a sorted tally, and a summary per group:

```r
patients |>                        # tally, most common first
  count(smoker) |>
  arrange(desc(n))

patients |>                        # one row per group
  group_by(arm) |>
  summarise(mean_age = mean(age))
```

## Making and changing columns

```{r, echo=FALSE}
crib(
  entry("mutate({your_data}, {new_name} = {calculation})",
        "Add a column built from the others. A comparison such as `sbp > 140` gives `TRUE`/`FALSE`.",
        "mutate(patients, high_bp = sbp > 140)", "M4"),
  entry('if_else({condition}, "{if_true}", "{if_false}")',
        "Picks between two values, row by row. Use it inside `mutate()`. With `tolower()`, which flattens case, it is the standard recoding pair.",
        'mutate(linelist, sex = if_else(tolower(sex) == "female", "Female", "Male"))', "M6"),
  entry("mutate({your_data}, {column} = as.numeric({column}))",
        "Convert text to numbers. Anything that was never a number becomes `NA`, with a warning.",
        "mutate(linelist, age = as.numeric(age))", "M6"),
  entry("is.na({column})",
        "`TRUE` where a value is missing. `!is.na({column})` is \"has a value\".",
        "is.na(weight)", "M6"),
  entry('factor({your_data}${column}, levels = c("{first}", "{second}", "{third}"))',
        "Fix the order of categories, instead of R's alphabetical default. This is what tidies your tables and plot axes. List every category, in the order you want.",
        'factor(patients$smoker, levels = c("Never", "Former", "Current"))', "M2"),
  entry('factor({your_data}${column}, levels = c("{lowest}", "{middle}", "{highest}"), ordered = TRUE)',
        "The same, plus permission to compare levels with `<` and `>`. Only needed for a real scale, like tumour stage.",
        NULL, "M2")
)
```

## Reshaping and joining tables

```{r, echo=FALSE}
crib(
  entry("{your_data} |> pivot_wider(names_from = {column}, values_from = {column})",
        "Long to wide: one column per visit. `names_from` names the column that supplies the new headings, `values_from` the one that fills the cells.",
        'labs |> filter(measure == "sbp") |> select(id, visit, value) |> pivot_wider(names_from = visit, values_from = value)', "M5"),
  entry('{your_data} |> pivot_longer(cols = c({column}, {column}), names_to = "{new_name}", values_to = "{new_name}")',
        "Wide back to long, the reverse move. The two new names are in quotes because those columns do not exist yet.",
        NULL, "M5"),
  entry('left_join({your_data}, {other_data}, by = "{column}")',
        "Bring columns from a second table across, matching on a column both tables share. Keeps every row of the first table.",
        'left_join(labs, patients, by = "id")', "M5"),
  entry('left_join({your_data}, select({other_data}, {column}, {column}), by = "{column}")',
        "Narrow the second table first when you only want a column or two. Keep the shared column in the `select()`.",
        'left_join(labs, select(patients, id, arm), by = "id")', "M5")
)
```

If a joined column comes back all `NA`, the key values did not match. That is the quiet failure to watch for.

## Plotting

Every plot is three parts: the data, the `aes()` mapping, and a `geom_` that draws it. The layers after the first are added with `+`.

```{r, echo=FALSE}
crib(
  entry("ggplot({your_data}, aes(x = {column}, y = {column})) + geom_point()",
        "Scatter plot: how two numbers relate.",
        "ggplot(patients, aes(x = bmi, y = sbp)) + geom_point()", "M7"),
  entry("aes(x = {column}, y = {column}, colour = {group})",
        "Map a third variable to colour. Inside `aes()` means \"from the data\", and ggplot adds the legend.",
        "aes(x = bmi, y = sbp, colour = arm)", "M7"),
  entry('geom_point(colour = "{colour_name}")',
        "A fixed colour, outside `aes()`, carrying no information.",
        'geom_point(colour = "steelblue")', "M7"),
  entry("ggplot({your_data}, aes(x = {group}, y = {column})) + geom_boxplot()",
        "One box per group: median line, middle half in the box.",
        "ggplot(patients, aes(x = arm, y = sbp)) + geom_boxplot()", "M7"),
  entry("ggplot({your_data}, aes(x = {column})) + geom_bar()",
        "Counts rows per category. No `y` needed, it counts for you.",
        "ggplot(patients, aes(x = arm)) + geom_bar()", "M7"),
  entry("ggplot({your_data}, aes(x = {column})) + geom_histogram(binwidth = {number})",
        "The shape of one numeric column. The {number} is the width of each bar, in the column's own units. Aim for about ten bars, then adjust.",
        "ggplot(patients, aes(x = age)) + geom_histogram(binwidth = 10)", "M10"),
  entry("ggplot({your_data}, aes(x = {time_column}, y = {column}, group = {id_column})) + geom_line()",
        "Joins each patient's readings into their own line. `group` says which rows belong to the same patient.",
        "ggplot(bp, aes(x = weeks, y = value, group = id)) + geom_line()", "M8"),
  entry("geom_line(linewidth = {number})",
        "Thicker lines, which survive being shrunk into a paper. Set outside `aes()`, because it is a fixed choice rather than something read from the data.",
        "geom_line(linewidth = 1)", "M8"),
  entry("geom_step()",
        "Draws a staircase rather than a sloped line, which is the correct shape for a survival curve.",
        NULL, "M13"),
  entry('geom_smooth(method = "lm")',
        "Adds a fitted straight line through the points, with a grey band for its 95% confidence interval. Add `se = FALSE` to leave the band off.",
        NULL, "M12"),
  entry("facet_wrap(~ {group})",
        "One panel per group, all on the same scale. Use when a single panel is too crowded.",
        "facet_wrap(~ arm)", "M8"),
  entry('labs(x = "{axis_title}", y = "{axis_title}", colour = "{legend_title}")',
        "Readable axis and legend titles. Any text you like, in quotes.",
        'labs(x = "Weeks", y = "Systolic BP", colour = "Treatment arm")', "M7, M8"),
  entry("theme_minimal()",
        "A cleaner look for a paper.",
        NULL, "M8"),
  entry('scale_colour_brewer(palette = "Dark2")',
        "Colours a reader with red-green colour blindness can still tell apart. Worth it from three groups up. Use `scale_fill_brewer()` for bars and boxes.",
        NULL, "M8"),
  entry("scale_colour_viridis_d()",
        "The same idea for an ordered variable; `scale_colour_viridis_c()` for a continuous one.",
        NULL, "M8")
)
```

## Tables and tests

```{r, echo=FALSE}
crib(
  entry('{your_data} |> summary_factorlist("{group}", c("{column}", "{column}"))',
        "The whole of Table 1 in one line: one column per {group}, one row per variable. Here the names go **in quotes**, unlike in `filter()`. Add as many variables inside `c()` as you need. From the `finalfit` package.",
        'patients |> summary_factorlist("arm", c("age", "sex"))', "M9"),
  entry('{your_data} |> summary_factorlist("{group}", c("{column}", "{column}"), p = TRUE)',
        "Adds a p-value per variable. One test per variable, not per level. Leave it off a randomised baseline table.",
        'patients |> summary_factorlist("sex", c("age", "bmi"), p = TRUE)', "M9"),
  entry("t.test({column} ~ {group}, data = {your_data})",
        "Compares the mean of a number between two groups. Read `~` as \"split by\".",
        "t.test(change ~ arm, data = bp_change)", "M11"),
  entry("chisq.test(table({your_data}${column}, {your_data}${column}))",
        "Tests whether two categories are associated.",
        "chisq.test(table(trial$arm, trial$response))", "M11"),
  entry("{result} |> tidy()",
        "Turns any test or model into a tidy data frame, with columns like `estimate` and `p.value`. The {result} is the whole test or model. From the `broom` package.",
        "t.test(change ~ arm, data = bp_change) |> tidy()", "M11")
)
```

`tidy()` returns up to ten columns, and the output panel hides the ones that do not fit behind a small arrow at the top right, so the confidence interval and p-value are often out of sight. Keep the ones you would report:

```r
t.test(change ~ arm, data = bp_change) |>
  tidy() |>
  select(estimate, conf.low, conf.high, p.value)
```

For a t-test, `estimate` is the first group's mean minus the second's, with the groups in alphabetical order, so check which way round it is before you report it. Lesson 11 explains every column.

### Reading `e-17` and the like

R writes very small numbers in **scientific notation**. `e-17` means "shift the decimal point 17 places left", and the exponent does all the work:

| R shows | Means | Report as |
|---|---|---|
| `6.78e-17` | 0.0000000000000000678 | p < 0.001 |
| `1.98e-03` | 0.00198 | p = 0.002 |
| `4.20e-02` | 0.042 | p = 0.042 |

`1.98e-03` is 0.002, **not** 1.98. A p-value can never exceed 1, so if yours looks bigger, you have missed the exponent. (M12)

## Models

All of them share the same formula shape: outcome, tilde, predictors. Add a predictor with `+`.

```{r, echo=FALSE}
crib(
  entry("lm({outcome} ~ {predictor}, data = {your_data})",
        "Linear regression, for a numeric outcome.",
        "lm(change ~ arm, data = bp_change)", "M12"),
  entry("lm({outcome} ~ {predictor} + {predictor}, data = {your_data})",
        "The same, adjusted for a second predictor. Adding predictors is what regression is for.",
        "lm(change ~ baseline + arm, data = bp_change)", "M12"),
  entry("glm({outcome} ~ {predictor}, data = {your_data}, family = binomial)",
        "Logistic regression, for a yes/no outcome.",
        "glm(response ~ arm, data = trial, family = binomial)", "M12"),
  entry("{model} |> tidy(exponentiate = TRUE, conf.int = TRUE)",
        "Odds ratios (or hazard ratios) with confidence intervals. Both are switches, and both are off by default.",
        "glm(response ~ arm, data = trial, family = binomial) |> tidy(exponentiate = TRUE, conf.int = TRUE)", "M12"),
  entry("Surv({your_data}${time_column}, {your_data}${event_column})",
        "Packages follow-up time and censoring into one outcome. Inside the model functions below, name the two columns without `{your_data}$`, because `data =` already says which table they come from.",
        "Surv(trial$time, trial$status)", "M13"),
  entry("survfit(Surv({time_column}, {event_column}) ~ {group}, data = {your_data})",
        "Kaplan-Meier estimate, one curve per group.",
        "survfit(Surv(time, status) ~ arm, data = trial)", "M13"),
  entry("survdiff(Surv({time_column}, {event_column}) ~ {group}, data = {your_data})",
        "The log-rank test: do the curves differ?",
        "survdiff(Surv(time, status) ~ arm, data = trial)", "M13"),
  entry("coxph(Surv({time_column}, {event_column}) ~ {group}, data = {your_data})",
        "Cox model, for the hazard ratio.",
        "coxph(Surv(time, status) ~ arm, data = trial)", "M13")
)
```

To draw the Kaplan-Meier curves, tidy the fit and step the line:

```r
km <- survfit(Surv(time, status) ~ arm, data = trial)

tidy(km) |>
  ggplot(aes(x = time, y = estimate, colour = strata)) +
  geom_step()
```

## Reporting

```{r, echo=FALSE}
crib(
  entry("{your_table} |> kable()",
        "Typesets a data frame as a proper table in a rendered document. Pipe a table into it as the last step.",
        'patients |> summary_factorlist("arm", c("age", "sex")) |> kable()', "M14"),
  entry("start_report()",
        "Puts the report template in your working directory and opens it.",
        NULL, "M14")
)
```

**Inline code** writes a number straight into a sentence, so your text can never disagree with your data. In a Quarto document, wrap the expression in backticks with an `r` at the front:

```{r, echo=FALSE, comment=""}
cat('The trial randomised `r nrow(patients)` patients.')
```

That comes out in the finished document as "The trial randomised 100 patients." (M14)

## Course housekeeping

```{r, echo=FALSE}
crib(
  entry("learn()",
        "List the lessons."),
  entry('learn("{lesson_id}")',
        "Open a lesson. The id is the one `learn()` lists, in quotes.",
        'learn("m4")'),
  entry("check_setup()",
        "Check your machine and install anything missing."),
  entry("use_local_library()",
        "Move your package library out of OneDrive, if installs keep failing."),
  lessons = FALSE
)
```

## Which lesson taught what

| Lesson | Tools |
|---|---|
| M2 Building blocks | `glimpse`, `head`, `class`, `table`, `factor` |
| M3 Wrangling I | `filter`, `select`, `arrange`, `desc`, `count`, the pipe |
| M4 Wrangling II | `mutate`, `group_by`, `summarise` |
| M5 Tidy data in practice | `pivot_wider`, `pivot_longer`, `left_join` |
| M6 Reading & cleaning | `read_csv`, `read_excel`, `if_else`, `tolower`, `as.numeric`, `is.na` |
| M7 Visualisation I | `ggplot`, `aes`, `geom_point`, `geom_boxplot`, `geom_bar`, `labs` |
| M8 Visualisation II | `geom_line`, `facet_wrap`, `theme_minimal`, `scale_colour_brewer` |
| M9 Summary tables | `summary_factorlist` |
| M10 Describing data | `summary`, `geom_histogram`, `mean`, `sd`, `median`, `IQR` |
| M11 Comparing groups | `t.test`, `chisq.test`, `tidy` |
| M12 Regression | `lm`, `glm`, `geom_smooth` |
| M13 Survival | `Surv`, `survfit`, `survdiff`, `coxph` |
| M14 Reproducible reports | `kable`, inline code, `start_report` |
