---
title: "R for Clinical Statistics: Wrangling Clinical Data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Your first hands-on lesson: pick rows, count groups, and sort a clinical dataset, one safe step at a time."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

# The teaching dataset ships with the package; see data-raw/patients.R.
# 100 synthetic trial patients, one row each. No real patient data.
data("patients", package = "r4clinstats")
```

## Welcome

This is your first time *doing* R rather than reading about it. A few promises before we start:

> **You cannot break anything.** Every box below is a safe sandbox. Run code, get it wrong, run it again. That is exactly how everyone learns this.
>
> **Red text is not failure.** Error messages are normal and expected, even for people who have used R for years. They are just R asking you to adjust something small.
>
> **You are never stuck.** Every exercise has a **Hint** button, and then a **Solution** button. Use them freely and without guilt.

In this lesson you'll learn to do three things to a real clinical dataset: **pick the rows you want**, **count groups**, and **sort**. That's the backbone of nearly every analysis you'll ever do.

## Meet your patients

We'll work with a dataset called `patients`, a (made-up) clinical trial with 100 people in it. Before touching it, one idea that will save you months of pain later.

This dataset is **tidy**, which means:

- **each row is one observation** (here, one patient), and
- **each column is one variable** (`age`, `sex`, the treatment `arm`, and so on).

That tidy shape is the whole reason the commands below will "just work." And here's the part most courses skip: **you decide this shape when you design your data-collection form, not when you sit down to analyse.** Sketch the table you'll need *before* you collect a single data point, working backwards from the analysis you're planning. We'll come back to this.

Enough talk. Press **Run Code** to look at the data:

```{r meet, exercise=TRUE}
glimpse(patients)
```

`glimpse()` gives you a sideways summary: every column, its type (`<chr>` = text, `<dbl>` = number), and the first few values. You can see all 100 patients are there, each described by 7 variables.

## The pipe: `|>`

Almost all modern R is written with the **pipe**, which looks like `|>`. Read it out loud as the words **"and then"**.

It takes whatever is on its left and feeds it into the function on its right. So this:

```{r pipe-demo, exercise=TRUE}
patients |>
  filter(age > 70)
```

reads as: *"take `patients`, **and then** keep only the rows where age is over 70."* Press **Run Code** and you'll get just the older patients.

That's it. That's the pipe. You'll use it constantly.

<div style="font-size: 0.9em; color: #666;">
*(You may see older code that uses `%>%` instead of `|>`. They do the same job; `|>` is the newer, built-in version.)*
</div>

## Picking rows with `filter()`

`filter()` keeps only the rows that match a condition. You just saw it with a number (`age > 70`). Now try it with **text**.

The trial has two treatment arms, `"Placebo"` and `"Active"`. **Fill in the blank** so the code keeps only the patients on the active drug, then press **Run Code** and **Submit Answer**:

```{r filter-arm, exercise=TRUE, exercise.blanks="___+"}
patients |>
  filter(arm == "____")
```

```{r filter-arm-hint}
# The treatment arms are written with a capital first letter.
# Look back at the data with glimpse(). Is the value "Active" or "active"?
```

```{r filter-arm-solution}
patients |>
  filter(arm == "Active")
```

```{r filter-arm-check}
grade_this({
  if (is.data.frame(.result) && nrow(.result) == 0) {
    fail("That ran, but it returned 0 patients. R is case-sensitive, so active and Active are different to it. Check the capital letter.")
  }
  pass_if_equal(message = "Exactly right. Those are the {nrow(.result)} patients on the active drug. Notice you matched text with a *double* `==`, not a single `=`.")
  fail("Close. Make sure you're keeping the `Active` arm. Try the Hint if you'd like a nudge.")
})
```

Notice the **double equals** `==`. In R, `==` means *"is it equal to?"* (a question), while a single `=` means *"set this value"* (a command). Mixing them up is a rite of passage; now you know.

## A quick word on *why* tidy matters

Before we count anything, look at what goes wrong when a column **isn't** clean. Imagine a study where "sex" was typed into a free-text box. The same single variable arrives looking like this:

```
sex:   "M"   "Male"   "male"   "1"   " m "   "Female"   "F"   "female"
```

Now you ask R to count how many of each:

```
# what you hoped for          # what you actually get
  Female   52                   "M"      18
  Male     48                   "Male"   14
                                "male"    9
                                "1"      12
                                " m "     4
                                "F"      28
                                "Female" 11
                                "female" 13
```

**One variable. Eight groups.** R is doing exactly what you asked, no more and no less. This is why, back at the design stage, you use a **dropdown with fixed options**, never a free-text box, for anything you'll want to count or group by. Cleaning up this mess afterwards is a whole skill of its own (we cover it later), but the cheapest fix is to never create the mess.

Our `patients` data was collected properly, so `sex` only ever contains `"Female"` or `"Male"`. Which means counting it is about to be wonderfully easy.

## Counting groups with `count()`

`count()` tells you how many rows fall into each group of a column. **Fill in the blank** to count how many patients are of each `sex`:

```{r count-sex, exercise=TRUE, exercise.blanks="___+"}
patients |>
  count(____)
```

```{r count-sex-hint}
# You want to count by the sex column.
# Type the column name exactly as it appears: sex
```

```{r count-sex-solution}
patients |>
  count(sex)
```

```{r count-sex-check}
grade_this({
  pass_if_equal(message = "Perfect: 43 female and 57 male patients, counted in one line. Now try re-running it with `arm` in place of `sex` to count the treatment groups instead.")
  fail("Not quite: pop the column name `sex` inside `count()`. The Hint has it if you need it.")
})
```

One line, and you have your group sizes. This little verb is the engine behind a great deal of clinical reporting.

## Sorting with `arrange()`

`arrange()` reorders the rows. By default it sorts smallest-to-largest. This time the blank is a little bigger. **Type the column you want to sort by** so the youngest patients appear first:

```{r arrange-age, exercise=TRUE, exercise.blanks="___+"}
patients |>
  arrange(____)
```

```{r arrange-age-hint}
# Which column holds each patient's age?
# It's called: age
```

```{r arrange-age-solution}
patients |>
  arrange(age)
```

```{r arrange-age-check}
grade_this({
  pass_if_equal(message = "There we go: sorted youngest first. (Tip for later: wrap the column in `desc()`, like `arrange(desc(age))`, to flip it to oldest first.)")
  fail("Almost: give `arrange()` the `age` column to sort on.")
})
```

## Choosing columns with `select()`

Last one, and you'll write a touch more this time. `filter()` picks *rows*; **`select()` picks *columns***. That's the only new idea here.

We want a smaller table showing just three columns: `id`, `age`, and `arm`. The verb is missing, so **fill in the function name**:

```{r select-cols, exercise=TRUE, exercise.blanks="___+"}
patients |>
  ____(id, age, arm)
```

```{r select-cols-hint}
# The verb for choosing columns is the opposite job to filter().
# It is called: select
```

```{r select-cols-solution}
patients |>
  select(id, age, arm)
```

```{r select-cols-check}
grade_this({
  pass_if_equal(message = "That's it: a tidy three-column table. You just chose rows AND columns, the two halves of pulling out exactly the data you want.")
  fail("Nearly! The column-picking verb is `select`. The Hint will confirm it.")
})
```

## Practice

Four verbs, no blanks. Build each answer yourself; the **Hint** and **Solution** have your back if you get stuck.

**1.** Show only the patients on the `Active` arm.

```{r pr-filter, exercise=TRUE}
# Your code here

```

```{r pr-filter-hint}
# filter() keeps matching rows. Match text with a double ==:
# patients |> filter(arm == "Active")
```

```{r pr-filter-solution}
patients |>
  filter(arm == "Active")
```

```{r pr-filter-check}
grade_this({
  pass_if_equal(message = "Spot on. {nrow(.result)} patients on the Active arm, picked out in one line.")
  fail("Not quite. Use `filter(arm == \"Active\")`, with a double `==` and a capital A.")
})
```

**2.** Count how many patients are in each `smoker` group, then sort from most to fewest.

```{r pr-count, exercise=TRUE}
# Your code here

```

```{r pr-count-hint}
# Two steps joined by the pipe. desc() sorts largest first:
# patients |> count(smoker) |> arrange(desc(n))
```

```{r pr-count-solution}
patients |>
  count(smoker) |>
  arrange(desc(n))
```

```{r pr-count-check}
grade_this({
  pass_if_equal(message = "That is the pattern: count to tally, then arrange to sort. `desc()` flips it to most-first.")
  fail("Close. Count by `smoker`, then `arrange(desc(n))` to sort from most to fewest. The Hint lays out both steps.")
})
```

**3.** In one pipeline, keep only patients older than 70, then show just their `id`, `age` and `arm`.

```{r pr-pipeline, exercise=TRUE}
# Your code here

```

```{r pr-pipeline-hint}
# filter() chooses rows, select() chooses columns. Chain them with the pipe:
# patients |> filter(age > 70) |> select(id, age, arm)
```

```{r pr-pipeline-solution}
patients |>
  filter(age > 70) |>
  select(id, age, arm)
```

```{r pr-pipeline-check}
grade_this({
  pass_if_equal(message = "A three-step pipeline, written from scratch. Filter to choose rows, then select to choose columns.")
  fail("Almost. Chain two steps with the pipe: `filter(age > 70)` and then `select(id, age, arm)`.")
})
```

## You did it 🎉

Look at what you can now do to a real clinical dataset, entirely in code:

- **`filter()`** keeps the rows that match a condition
- **`count()`** tallies how many fall into each group
- **`arrange()`** sorts the rows
- **`select()`** keeps the columns you want
- **`|>`** chains steps together, reading "and then"

Every one of these worked first or second time, and that feeling of *"oh, I can actually do this"* is the whole point of today. The fancier analyses later are just more of these small, learnable steps stacked together.

And the bigger lesson underneath it all: this was easy **because the data was tidy**, and the data was tidy **because someone designed the collection form with the analysis in mind.** That's the habit worth carrying into your own projects.
