---
title: "R for Clinical Statistics: The Building Blocks of R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Before you wrangle: what an object is, the handful of value types you'll meet, and why type matters."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
```

## Welcome

The same three promises as ever, because they don't stop being true:

> **You cannot break anything.** Every box is a safe sandbox. Run it, get it wrong, run it again.
>
> **Red text is not failure.** An error message is R asking for a small adjustment, nothing more.
>
> **You are never stuck.** Every exercise has a **Hint**, then a **Solution**. Use them freely.

This lesson is the vocabulary the rest of the course leans on. By the end you'll know what an **object** is, the few **types** of value R cares about, and why getting the type right saves you grief later. It's gentle on purpose, mostly run-and-see.

## Meet your patients

Here's the dataset we'll use throughout. Press **Run Code**:

```{r glimpse-meet, exercise=TRUE}
glimpse(patients)
```

That's a **data frame**: a stack of columns, each one a single kind of value, lined up so every row is one patient. `glimpse()` shows each column's type in angle brackets: `<chr>` for text, `<dbl>` and `<int>` for numbers. Those types are what this lesson is about.

## Objects: giving things names

In R you do almost everything by putting a value into a named **object** with the arrow `<-`. Run this:

```{r assign-demo, exercise=TRUE}
my_value <- 42
my_value
```

The first line says *"let `my_value` be 42"*; the second line shows it. Once a value has a name, you can use it again and again.

Your turn. **Fill in any number**, then run:

```{r assign-weight, exercise=TRUE, exercise.blanks="___+"}
weight <- ____
weight
```

```{r assign-weight-hint}
# Pick any number and put it after the arrow, for example:
# weight <- 70
```

```{r assign-weight-solution}
weight <- 70
weight
```

```{r assign-weight-check}
grade_this({
  if (is.numeric(.result) && length(.result) == 1) {
    pass("That's an object. `weight` now holds {.result}, ready to reuse.")
  }
  fail("Put a single number after the arrow, for example `weight <- 70`.")
})
```

```{r assign-weight-wrong, eval=FALSE, include=FALSE}
# expect: single number
weight <- "70"
weight
```

## What *type* is it?

Every column has a **type**, and R treats each type differently. The commonest are:

- **character**, text like `id` or `sex` (shown as `<chr>`)
- **numeric**, numbers like `age` or `sbp` (`<dbl>` or `<int>`)
- **logical**, `TRUE` / `FALSE` answers
- **factor**, a category with a fixed set of options (more in a moment)
- and **`NA`**, meaning *missing*, which can sit inside any of them.

Ask R the type of a column with `class()`. Run this:

```{r class-demo, exercise=TRUE}
class(patients$age)
```

The `patients$age` part means *"the `age` column of `patients`"*, and R answers `"numeric"`. Now you: **fill in the column name** to check the type of `sex`:

```{r class-col, exercise=TRUE, exercise.blanks="___+"}
class(patients$____)
```

```{r class-col-hint}
# Type the column name exactly as it appears in glimpse(patients): sex
```

```{r class-col-solution}
class(patients$sex)
```

```{r class-col-check}
grade_this({
  if (identical(.result, "NULL")) {
    fail("That column name isn't in `patients`. Check the spelling against glimpse(patients).")
  }
  pass_if_equal(message = "Yes, `sex` is stored as text, which R labels `chr`, short for character.")
})
```

```{r class-col-wrong, eval=FALSE, include=FALSE}
# expect: column name
class(patients$sexx)
```

## Factors: categories with an order

Some categories have a natural sequence. `smoker` is one: **Never**, then **Former**, then **Current**. Stored as plain text, R has no way of knowing that. Press **Run Code** to see what it does instead:

```{r smoker-alpha, exercise=TRUE}
table(patients$smoker)
```

Alphabetical: Current, Former, Never. Every table you build, every plot axis and every model you fit will inherit that order, and it reads backwards to any clinician. A **factor** fixes it by recording the categories *and* the order you want them in. Run this:

```{r smoker-levels, exercise=TRUE}
smoker_f <- factor(patients$smoker, levels = c("Never", "Former", "Current"))

table(smoker_f)
```

Same counts, sensible order. That is `levels =` doing the work.

**Fill in the last level** so the order runs Never, Former, Current:

```{r factor-smoker, exercise=TRUE, exercise.blanks="___+"}
factor(patients$smoker,
       levels = c("Never", "Former", "____"),
       ordered = TRUE)
```

```{r factor-smoker-hint}
# The three levels in order are Never, Former and Current.
# Fill in the last one, spelled exactly as in the data.
```

```{r factor-smoker-solution}
factor(patients$smoker,
       levels = c("Never", "Former", "Current"),
       ordered = TRUE)
```

```{r factor-smoker-check}
grade_this({
  if (any(is.na(.result))) {
    fail("Almost: some values became NA, which means a level name doesn't match the data exactly. Check the capitals: it should be Current.")
  }
  pass_if_equal(message = "Perfect. `smoker` is now an *ordered* factor: Never < Former < Current. R now knows these run in a sequence.")
})
```

```{r factor-smoker-wrong, eval=FALSE, include=FALSE}
# expect: capitals
factor(patients$smoker,
       levels = c("Never", "Former", "current"),
       ordered = TRUE)
```

Look at the bottom of the output: `Levels: Never < Former < Current`. Those `<` signs are what `ordered = TRUE` added, and it is worth being exact about what they buy you.

### What `ordered = TRUE` actually changes

`levels =` sets the order, and that alone is what fixes your tables, your plot axes and the reference level your models compare against. `ordered = TRUE` adds one further thing: permission to *compare* levels with `<` and `>`. Run this and read the two answers:

```{r ordered-compare, exercise=TRUE}
smoker_plain <- factor(patients$smoker, levels = c("Never", "Former", "Current"))
smoker_ord   <- factor(patients$smoker, levels = c("Never", "Former", "Current"), ordered = TRUE)

head(smoker_plain > "Never")   # R refuses, and warns you
head(smoker_ord   > "Never")   # a straight answer: has this patient ever smoked?
```

The plain factor returns `NA` for every patient with a warning that `>` is "not meaningful". The ordered one answers the question.

So for most day-to-day clinical work, `levels =` is the part you need. Reach for `ordered = TRUE` when the categories are a real scale that you will compare or model as one: tumour stage, NYHA class, a Likert response.

## Yes / no answers: logical values

Ask a question of a column and R answers `TRUE` or `FALSE` for every row. **Fill in the age cut-off** to flag patients older than 65:

```{r logical-age, exercise=TRUE, exercise.blanks="___+"}
patients$age > ____
```

```{r logical-age-hint}
# Type the cut-off as a plain number, with no quotes: 65
```

```{r logical-age-solution}
patients$age > 65
```

```{r logical-age-check}
grade_this({
  pass_if_equal(message = "There you go: a column of TRUE/FALSE answers, one per patient. That's a *logical* vector, and it's what `filter()` quietly uses (your next lesson).")
  fail("Not quite: we want patients older than 65, so the number is 65, with no quotes.")
})
```

## Why the type matters

Get the type wrong and ordinary things stop working. Picture an age that was typed into a text box, so it arrived as **text** rather than a **number**:

```
age_text <- "65"     # note the quotes: this is text, not a number

age_text + 1
#> Error: non-numeric argument to binary operator
```

R can't add 1 to the word "65". The cure is to collect `age` as a number in the first place, the same design point from the tidy-data lesson. When something later refuses to add up or sort properly, the type is the first thing to check.

## Practice

Three to finish, and no blanks this time. Write each one yourself. The **Hint** and **Solution** are there if you want them.

**1.** Create an object called `x` that holds the value `10`, then print it.

```{r make-x, exercise=TRUE}
# Your code here

```

```{r make-x-hint}
# Put 10 into x with the arrow, then type x on its own line to show it:
# x <- 10
# x
```

```{r make-x-solution}
x <- 10
x
```

```{r make-x-check}
grade_this({
  env <- .envir_result
  if (exists("x", envir = env, inherits = FALSE)) {
    val <- get("x", envir = env)
    if (is.numeric(val) && length(val) == 1 && val == 10) {
      pass("That's it. `x` now holds 10, and you can reuse it anywhere from here on.")
    }
    fail("You've made `x`, but it should hold the number 10. Try `x <- 10`.")
  }
  fail("I can't find an object called `x`. Create one with `x <- 10`, then type `x` on the next line to print it.")
})
```

**2.** Create an object called `size` that holds the three words `"small"`, `"medium"` and `"large"`, in that order.

```{r make-size, exercise=TRUE}
# Your code here

```

```{r make-size-hint}
# Collect several values with c(), keeping each word in quotes:
# size <- c("small", "medium", "large")
```

```{r make-size-solution}
size <- c("small", "medium", "large")
size
```

```{r make-size-check}
grade_this({
  env <- .envir_result
  if (!exists("size", envir = env, inherits = FALSE)) {
    fail("I can't find an object called `size`. Build it with `size <- c(\"small\", \"medium\", \"large\")`.")
  }
  val <- get("size", envir = env)
  if (identical(val, c("small", "medium", "large"))) {
    pass("Three text values, in order, under one name. That is a character vector.")
  }
  fail("Close. `size` should be the three words small, medium and large, in that order, each in quotes.")
})
```

**3.** Now turn `size` into an **ordered factor** that runs from `small` up to `large`.

```{r mk-factor-setup}
size <- c("small", "medium", "large")
```

```{r mk-factor, exercise=TRUE}
# size already exists. Make it an ordered factor:

```

```{r mk-factor-hint}
# Give factor() the levels in the order you want, and switch ordering on:
# factor(size, levels = c("small", "medium", "large"), ordered = TRUE)
```

```{r mk-factor-solution}
factor(size, levels = c("small", "medium", "large"), ordered = TRUE)
```

```{r mk-factor-check}
grade_this({
  if (!is.factor(.result)) {
    fail("The result should be a factor. Wrap `size` in `factor(...)`.")
  }
  if (!is.ordered(.result)) {
    fail("Almost there. Add `ordered = TRUE` so the sizes have a direction, from small up to large.")
  }
  if (!identical(levels(.result), c("small", "medium", "large"))) {
    fail("Check the level order. It should run small, then medium, then large.")
  }
  pass("An ordered factor: small < medium < large. R now treats the sizes as a sequence, not three loose labels.")
})
```

## Recap

You now have the words the rest of the course uses without stopping to explain:

- an **object** holds a value under a name (`<-`)
- a **data frame** is columns of values, one type each, lined up in rows
- the types you'll meet are **character**, **numeric**, **logical** and **factor**, with **`NA`** for missing
- `class()` tells you which one you've got

None of the later lessons assume more than this. With the vocabulary in place, you're ready to start wrangling.
