---
title: "R for Clinical Statistics: Describing Data"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: "Choosing the right summary: histograms first, mean (SD) for symmetric data, median (IQR) when the shape is lopsided."
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(dplyr)
library(ggplot2)

gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

data("patients", package = "r4clinstats")
data("outcomes", package = "r4clinstats")
```

## Welcome

Part 3 begins here: the analysis lessons. Before testing anything, you need to describe it, and describing a variable well means choosing the right summary. This lesson gives you the rule for choosing, and the habit that makes the rule easy: **look at the shape first**.

The promises hold: nothing breaks, red text is normal, and **Hint** then **Solution** are a click away.

## Meet the trial outcomes

A new table joins the course. `outcomes` records what happened to each of the 100 patients over a year of follow-up. Press **Run Code**:

```{r meet-outcomes, exercise=TRUE}
glimpse(outcomes)
```

Four facts per patient: whether they responded to treatment (`response`), whether they had an adverse event (`adverse_event`), how many days they were followed (`time`), and whether the event being watched for actually happened (`status`, where 1 means it did). The `time` and `status` pair belongs to survival analysis, which gets its own lesson; today they make a useful example of a badly behaved variable.

## All the numbers at once

`summary()` is the quickest look at a numeric variable. **Run this** on the patients' ages:

```{r summary-age, exercise=TRUE}
summary(patients$age)
```

Six numbers: the smallest and largest, the quartiles, the median and the mean. Notice the mean (62.2) and the median (63) nearly agree. Keep that in mind; it will matter in a moment.

## Look at the shape first

A histogram shows a variable's whole shape: it chops the number line into bins and counts how many patients land in each. **Fill in the geom** that does this:

```{r hist-age, exercise=TRUE, exercise.blanks="___+"}
ggplot(patients, aes(x = age)) +
  ____(binwidth = 5)
```

```{r hist-age-hint}
# The geom that bins a numeric variable and counts each bin is:
# geom_histogram
```

```{r hist-age-solution}
ggplot(patients, aes(x = age)) +
  geom_histogram(binwidth = 5)
```

```{r hist-age-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Keep the `ggplot(...)` start and add a geom.")
  }
  stats <- vapply(.result$layers, function(l) class(l$stat)[1], character(1))
  if ("StatBin" %in% stats) {
    pass("A single hill, roughly symmetric, centred in the early 60s. When a variable looks like this, the mean and SD describe it faithfully.")
  }
  fail("Almost: `geom_histogram()` bins the ages and counts each bin. The Hint has it.")
})
```

Age is a tidy hill. That shape is what makes the mean trustworthy here.

## Mean and standard deviation

For a symmetric variable, report the **mean** with the **standard deviation** (SD), which measures how spread out the values are. **Fill in the function** for the SD:

```{r mean-sd, exercise=TRUE, exercise.blanks="___+"}
patients |>
  summarise(mean_age = mean(age), sd_age = ____(age))
```

```{r mean-sd-hint}
# The standard deviation function is called:
# sd
```

```{r mean-sd-solution}
patients |>
  summarise(mean_age = mean(age), sd_age = sd(age))
```

```{r mean-sd-check}
grade_this({
  pass_if_equal(message = "Age is 62.2 years (SD 10.9). Written that way, mean (SD), it is exactly what a paper's Table 1 reports for a symmetric variable.")
  fail("Almost: the standard deviation function is `sd`.")
})
```

## When the mean misleads

Now the badly behaved variable. **Run this** histogram of follow-up time:

```{r hist-time, exercise=TRUE}
ggplot(outcomes, aes(x = time)) +
  geom_histogram(binwidth = 20)
```

No tidy hill this time: events scattered through the year, then a tall spike at 365, the 58 patients who reached one year with no event at all. The shape is lopsided, and watch what that does to the summaries. **Fill in the function** for the median:

```{r mean-median, exercise=TRUE, exercise.blanks="___+"}
outcomes |>
  summarise(mean_time = mean(time), median_time = ____(time))
```

```{r mean-median-hint}
# The middle value of a variable is its:
# median
```

```{r mean-median-solution}
outcomes |>
  summarise(mean_time = mean(time), median_time = median(time))
```

```{r mean-median-check}
grade_this({
  pass_if_equal(message = "The mean says 282 days; the median says 365. They disagree because the shape is lopsided, and when they disagree, the median with the interquartile range (IQR) is the honest pair. (Summaries of a censored follow-up time mislead in deeper ways too; the survival lesson has the proper tool.)")
  fail("Almost: the middle value is the `median`.")
})
```

```{r mean-median-wrong, eval=FALSE, include=FALSE}
# expect: median
outcomes |>
  summarise(mean_time = mean(time), median_time = mean(time))
```

The rule, then: **plot the shape first**. A symmetric hill earns mean (SD); anything lopsided gets median (IQR), which you compute with `median()` and `IQR()`.

## Summaries by group

Descriptions usually come per group. This is `group_by()` and `summarise()` from the wrangling lessons, now with two summaries at once. **Fill in the grouping column** to describe blood pressure per treatment arm:

```{r group-describe, exercise=TRUE, exercise.blanks="___+"}
patients |>
  group_by(____) |>
  summarise(mean_sbp = mean(sbp), sd_sbp = sd(sbp))
```

```{r group-describe-hint}
# One row of summaries per treatment arm, so group by:
# arm
```

```{r group-describe-solution}
patients |>
  group_by(arm) |>
  summarise(mean_sbp = mean(sbp), sd_sbp = sd(sbp))
```

```{r group-describe-check}
grade_this({
  pass_if_equal(message = "139.3 (14.7) against 139.6 (15.7): the arms start from the same blood pressure, which is exactly what randomisation is for. This is also what finalfit computed for you in the tables lesson; now you can see the machinery.")
  fail("Almost: group by `arm` so each treatment group gets its own row of summaries.")
})
```

## Practice

Three to finish, no blanks. Write each one yourself; **Hint** and **Solution** are there if you want them.

**1.** Draw a histogram of `bmi` from `patients`, with a `binwidth` of 2.

```{r pr-hist, exercise=TRUE}
# Your code here

```

```{r pr-hist-hint}
# ggplot(patients, aes(x = bmi)) + geom_histogram(binwidth = 2)
```

```{r pr-hist-solution}
ggplot(patients, aes(x = bmi)) +
  geom_histogram(binwidth = 2)
```

```{r pr-hist-check}
grade_this({
  if (!inherits(.result, "ggplot")) {
    fail("That should draw a plot. Start with `ggplot(patients, aes(...))`.")
  }
  stats <- vapply(.result$layers, function(l) class(l$stat)[1], character(1))
  if (!("StatBin" %in% stats)) {
    fail("Use `geom_histogram()` to bin and count the BMI values.")
  }
  maps <- c(list(.result$mapping), lapply(.result$layers, function(l) l$mapping))
  xs <- unlist(lapply(maps, function(m) {
    if (is.null(m$x)) NULL else rlang::as_label(m$x)
  }))
  if (!("bmi" %in% xs)) {
    fail("Put `bmi` on the x-axis inside `aes()`.")
  }
  pass("The shape check, written from scratch. BMI is another rough hill, so mean (SD) will serve it well.")
})
```

**2.** For each `smoker` group, report the mean and SD of `bmi`. Call the columns `mean_bmi` and `sd_bmi`.

```{r pr-group, exercise=TRUE}
# Your code here

```

```{r pr-group-hint}
# Group, then summarise twice in one call:
# patients |> group_by(smoker) |> summarise(mean_bmi = mean(bmi), sd_bmi = sd(bmi))
```

```{r pr-group-solution}
patients |>
  group_by(smoker) |>
  summarise(mean_bmi = mean(bmi), sd_bmi = sd(bmi))
```

```{r pr-group-check}
grade_this({
  pass_if_equal(message = "Mean (SD) of BMI for each smoking group, one line of code. That is a full row of a Table 1.")
  fail("Close. Group by `smoker`, then `summarise(mean_bmi = mean(bmi), sd_bmi = sd(bmi))`, keeping those column names.")
})
```

**3.** `time` is lopsided, so give it the right pair: the median and the IQR, from `outcomes`. Call them `median_time` and `iqr_time`.

```{r pr-median, exercise=TRUE}
# Your code here

```

```{r pr-median-hint}
# median() and IQR() inside one summarise:
# outcomes |> summarise(median_time = median(time), iqr_time = IQR(time))
```

```{r pr-median-solution}
outcomes |>
  summarise(median_time = median(time), iqr_time = IQR(time))
```

```{r pr-median-check}
grade_this({
  pass_if_equal(message = "Median 365 days, IQR 192.5. For a lopsided variable this pair tells the truth that the mean cannot.")
  fail("Close. Use `median(time)` and `IQR(time)` inside `summarise()`, with the names `median_time` and `iqr_time`.")
})
```

## Recap

Describing a variable is a two-step habit:

- **look at the shape first**, with `geom_histogram()`
- a symmetric hill gets **mean (SD)**; a lopsided shape gets **median (IQR)**
- **`group_by()` + `summarise()`** turns either into a per-group description

Next you stop describing differences and start testing them: is the gap between two groups more than chance could produce?
